---
title: "Data manipulation in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE, context="server"}
library(learnr)
library(gradethis)
library(glue)
library(dplyr)
library(curl)
library(jsonlite)
library(openssl)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

req <- curl_fetch_memory("https://api.ipify.org?format=json")

HASH <- as.character(md5(fromJSON(rawToChar(req$content))[[1]]))

event_recorder <- function(tutorial_id, 
                           tutorial_version,
                           user_id,
                           event, 
                           data) {
  REPO <- "/vol/home-vol2/dig/kuzilekj/ML4ED_2025/04_data_manipulation_in_r"
  
  if (file.exists(glue("{REPO}/data_{HASH}.rds"))) {
    t <- readRDS(glue("{REPO}/data_{HASH}.rds"))
  } else {
    t <- 
      tibble(
        time = .POSIXct(numeric(0)),
        tutorial_id = character(),
        tutorial_version = character(),
        user_id = character(),
        hash = character(),
        event = character(),
        data = list()
      )
  }
  
  t <- bind_rows(t, 
                 tibble(
                   time = Sys.time(),
                   tutorial_id = tutorial_id,
                   tutorial_version = tutorial_version,
                   user_id = user_id,
                   hash = HASH,
                   event = event,
                   data = list(data)
                 )
  )
  
  saveRDS(t, file = glue("{REPO}/data_{HASH}.rds"))
}

options(tutorial.event_recorder = event_recorder)

lecture_grade_code <- function() {
  grade_this_code(
    correct = "Correct!",
    incorrect = "Try again. {code_feedback()}"
  )
}
```

## Introduction
At the end of this lecture you will:

- learn how to use pipes for improved code readability
- find out alternative to `apply` family 
- learn the basics of data wrangling
- explore the basics of advanced plotting in R
  
> For simplicity, we will not be using the R and RStudio installed on your computer. We will be
using interactive excercises created using [learnr](https://cran.r-project.org/web/packages/learnr/index.html) package.

> When you find `___` in the code chunk, it means that you need to fill in the missing part of the code.

## Informed Consent
This informed consent form is for a project investigating the interactions with the online
tutorial platform created using `learnr` package.

Name of Investigator: Dr. Jakub Kuzilek (jakub.kuzilek@hu-berlin.de)

Name of Supervisor: Prof. Dr. Niels Pinkwart (pinkwart@hu-berlin.de)

Name of Institution: Humboldt-Universität zu Berlin

**You can download this informed consent in the Moodle course.**

**The data from multiple tutorials will be collected and analysed.The informed consent is the same. For each session, you will get hash, which is the only unique identifier of your interactions with the material.** 

### Information sheet

#### Introduction of investigators
Jakub Kuzilek is a researcher in fields of educational data mining and learning analytics since completion of his PhD in Artificial Intelligence and Biocybernetics at the Czech Technical University. In 2020 he joined Prof. Dr. Pinkwarts's research group and currently working on the project IMPACT.

#### Purpose of the study
The purpose of this study is to investigate the interactions with the online tutorial platform created using the `learnr` package. The data will be used to improve the design of the tutorial and to understand how users interact with the platform.

#### Type of Research Intervention
To support and investigate these learning and interaction processes interaction data will be used. In addition performance data of the students will be collected.

#### Research Data Management
The collected data will be anonymized and merged. The collected data are stored on the servers of the Computer and Media Service (CMS) of the Humboldt-Universität zu Berlin in Germany and the computers of the researchers. The data access is protected and regulated via a login and the access is restricted to the members of the project team. The personal data will be stored on secured location accessible via authenticated encrypted connection (VPN and SSH). The data are archived for another 10 years after the end of the project according to good scientific practice. The anonymized data may be published to enable Open Access for the general public. The participant may gain access to his/her recorded data, request a correction to his/her personal data record, or withdraw from the research study by contacting the Investigator:
Jakub Kuzilek at jakub.kuzilek@hu-berlin.de

In case of doubt as to the lawfulness of the processing of personal data, data subjects have the right to lodge a complaint with a Member State supervisory authority. The competent supervisory authority for Humboldt-Universität zu Berlin is Die Berliner Beauftragte für Datenschutz und Informationsfreiheit, Friedrichstraße 219, 10969 Berlin. The official data protection officer for Humboldt-Universität zu Berlin is: Gesine Hoffmann-Holland, Tel: +49 (30) 2093-20022, datenschutz@uv.hu-berlin.de.

#### Voluntary participation
Participation in this research project is entirely voluntary. It is up to the participant to decide whether or not to participate and there is no explicit or implicit obligation to participate. The participant has the possibility to cancel the participation at any time - even after the signing of the Informed Consent - without any disadvantages for the participant. 

#### Documentation of the consent
One copy of this document will be kept separately from the research records of this study using a pseudonym id in order to remove data after the request from the participant. The contact details of the participants will be deleted after the project is finished. To assure data deletion of the requesting subject a unique identifier in form of the email address and is required.

### Certificate of consent
- I agree to the collection of my pseudonymized interaction data in the course "Machine Learning for Education".
- I agree to the collection of my pseudonymized perfomance data in the course "Machine Learning for Education".
- I agree to have the anonymized interaction data and the anonymized performance data published under the Creative Commons Attribution 4.0 International (CC-BY) Licence (https://creativecommons.org/licenses/ by/4.0/) and so I allow it to be used by other people for other projects. 
- I agree to have the interaction data, and the performance data archived for 10 years after finishing of the project.

This hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:

**`r HASH`**

Please, keep it for your records in case you change your mind and want to delete your data.

```{r informed-consent, echo=FALSE}
question_radio("I agree to the collection of my pseudonymized interaction data in the course 'Machine Learning for Education'.", 
               correct = NULL, 
               answer("yes", correct = TRUE),
               answer("no", correct = TRUE)
)
```


## Advanced R programming
We will introduce two topics, which can be considered as advanced, but 
at the end both approaches simplify the programming in R. The first topic are 
`pipes`, which works in similar way to the pipes in the Linux bash command line. 
They enable chaining commands (functions) in more readable format with only a little
overhead compared to the normal function call. The second topic is package
`purrr`, which is the alternative to the apply function family providing more consistent
way for handling the iteration using functional programming paradigm.

## Pipes

Pipes are the way how to chain the function calls together. The main idea is to
pass the result of one function to the next function as an argument. There are two
types of pipes in R. The first one is the native pipe, which is available in R
version 4.1.0 and higher. The second one is the pipe operator `%>%`, which
comes from the `magrittr` package. It is also available in the `dplyr` package, 
which is part of the `tidyverse`. 

### Example 
Let us start with the example. Imagine you are writing the program for cooking pasta, 
each step is represented by one function. The input to the first function `get_pasta`
would be `type` of pasta and `number_of_people`, who are going to eat that pasta. The output
is a list containing the type of pasta and number of people. Try to define the function:

```{r pipes-get-pasta-example, exercise=TRUE}
___ <- function(___, ___) {
  print(paste0("Getting ",___," for ",___," persons."))
  list(___, ___)
}
```

```{r pipes-get-pasta-example-solution}
get_pasta <- function(type, number_of_people) {
  print(paste0("Getting ",type," for ",number_of_people," persons."))
  list(type, number_of_people)
}
```

```{r pipes-get-pasta-example-check}
lecture_grade_code()
```

Now we can define functions `cook_pasta` and `serve_pasta`. The first one will take the
pasta list and print the message about cooking. It will also `append` the cooking time
to the list and returns it. The second one will take the pasta list and print the message about serving
the pasta. It will return the same list as an output. Try to define the functions:
```{r pipes-cook-serve-pasta-example, exercise=TRUE}
___ <- function(___) {
  print(paste0("Cooking ", ___[[1]],".")) # get the type of pasta (1st element of the list)
  ___(___, "10 minutes")
}

___ <- function(___) {
  # get the type of pasta (1st element of the list),
  # number of people (2nd element of the list), 
  # cooking time (3rd element of the list)
  print(paste0("Serving ", ___[[___]], " for ", ___[[___]], "persons cooked for ", ___[[___]],"."))
  ___
}
```

```{r pipes-cook-serve-pasta-example-solution}
cook_pasta <- function(pasta) {
  print(paste0("Cooking ", pasta[[1]], "."))
  pasta[[3]] <- "10 minutes"
  pasta
}

serve_pasta <- function(pasta) {
  # get the type of pasta (1st element of the list),
  # number of people (2nd element of the list), 
  # cooking time (3rd element of the list)
  print(paste0("Serving ", pasta[[1]], " for ", pasta[[2]], " persons cooked for ", pasta[[3]], "."))
  pasta
}
```

```{r pipes-cook-serve-pasta-example-check}
lecture_grade_code()
```

### Cooking using variables
Normally, we will call the functions in defined order and store the interim results
in variables. Those variables can be used later in the code. Try this in the following
exercise. Get the `spaghetti` pasta for 20 people, cook it and serve it:
```{r pipes-pasta}
get_pasta <- function(type, number_of_people) {
  print(paste0("Getting ",type," for ",number_of_people," persons."))
  list(type, number_of_people)
}

cook_pasta <- function(pasta) {
  print(paste0("Cooking ", pasta[[1]], "."))
  pasta[[3]] <- "10 minutes"
  pasta
}

serve_pasta <- function(pasta) {
  # get the type of pasta (1st element of the list),
  # number of people (2nd element of the list), 
  # cooking time (3rd element of the list)
  print(paste0("Serving ", pasta[[1]], " for ", pasta[[2]], " persons cooked for ", pasta[[3]], "."))
  pasta
}
```

```{r pipes-objects-example, exercise=TRUE, exercise.setup = "pipes-pasta"}
pasta <- ___("___", ___)
cooked_pasta <- ___(pasta)
result <- ___(cooked_pasta)
```

```{r pipes-objects-example-solution}
pasta <- get_pasta("spaghetti", 20)
cooked_pasta <- cook_pasta(pasta)
result <- serve_pasta(cooked_pasta)
```

```{r pipes-objects-example-check}
lecture_grade_code()
```

### Cooking compressed
We can call function from function to pass the interim results as an argument. This 
will save some space and typing. Try to do the same as in the previous exercise:
```{r pipes-compressed-example, exercise=TRUE, exercise.setup = "pipes-pasta"}
result <- ___(___(___("___", 20)))
```

```{r pipes-compressed-example-solution}
result <- serve_pasta(cook_pasta(get_pasta("spaghetti", 20)))
```

```{r pipes-compressed-example-check}
lecture_grade_code()
```

### Cooking using pipes
The previous two examples are the standard way how to call functions in R. Both ways
are correct and they are used in the R code. But they are not very readable and
are sensitive for typing errors and somewhat are hard to read. Especially the 
second case. For hard readable code we got Lisp :) Luckily, we can use the pipes
to help us. We will be using base R pipe `|>` operator in our examples. Try to 
rewrite the previous example using the pipe:

```{r pipes-base-pipe-example, exercise=TRUE, exercise.setup = "pipes-pasta"}
result <-
  ___("spaghetti", 20) |> 
  ___() ___
  ___()
```

```{r pipes-base-pipe-example-solution}
result <-
  get_pasta("spaghetti", 20) |>
  cook_pasta() |>
  serve_pasta()
```

```{r pipes-base-pipe-example-check}
lecture_grade_code()
```

This form of writing the code helps you focus on verbs (=function names) and you
can read the function composition as a set of imperative actions. Not all the times 
using a pipe is useful, but in the most cases it improves the code readability and 
helps you with the sequences of linear operations. 

## Advanced iterations

The `purrr` package comes as an alternative to the `apply` function family. In 
most cases it provides the same functionality and performance like `apply` functions,
but it comes with more consistent function calls and names, thus it is easier to 
understand and work with. 

```{r purrr-setup, include=FALSE}
library(purrr)
```

### Example data
Before we start using the `map*` functions from `purrr`. We need to create example `data.frame` `df`
with 4 columns (`a`,`b`,`c`,`d`) and `10` rows. The columns will be filled with random numbers from normal
distribution using `rnorm` function. Create the `df` data.frame:
```{r purrr-data-example, exercise=TRUE}
___ <- ___(
  ___ = rnorm(___),
  ___,
  ___,
  ___
)
```

```{r purrr-data-example-solution}
df <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

```{r purrr-data-example-check}
lecture_grade_code()
```

### Apply refresher
You are familiar with the `apply` function family from the previous lecture. Use it
on the `df` `data.frame` to calculate the `mean` of each column (`MARGIN=2`, second argument):
```{r apply-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
___(___,2,___)
```

```{r apply-example-solution}
apply(df, 2, mean)
```

```{r apply-example-check}
lecture_grade_code()
```

### map
The `map` family of functions is the alternative to the `apply` function family. To call
the base function `map` you need to provide the first argument, which is the input
data (e.g. `data.frame`), the second argument is the function you want to apply.
Try to rewrite the previous example:
```{r map-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
___(___, ___)
```

```{r map-example-solution}
map(df, mean)
```

```{r map-example-check}
lecture_grade_code()
```

### map family
The `map` function is the most general one. It can be used with any type of input
data. It returns the result as a list. If one wants to have the result in a specific
format, the only thing you need to do is to add the suffix to the function name. 
In our case we want to have the result in a numeric vector format thus we can
add `_dbl` suffix to the function name. Try to modify previous example:
```{r map-dbl-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
map(df, mean)
```

```{r map-dbl-example-solution}
map_dbl(df, mean)
```

```{r map-dbl-example-check}
lecture_grade_code()
```

From the function name we know the output and the call is consistent
with other functions from the "family".The other fuctions from family are: 
`map_lgl`, `map_int`, `map_chr`.

### Own functions
Both `apply` and `map` family can work with your own defined functions. Define your
own function `my_mean` using `sum` and `length` functions and use it in the call
of `map_dbl` function:
```{r map-own-function-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
___ <- function(x) {
  ___(x)/___(x)
}

map_dbl(df, ___)
```

```{r map-own-function-example-solution}
my_mean <- function(x) {
  sum(x)/length(x)
}

map_dbl(df, my_mean)
```

```{r map-own-function-example-check}
lecture_grade_code()
```

### Anonymous functions
Within the call of the `map` function you can also use the anonymous functions. This 
is handy when the function is used only once and you do not want to define it
explicitly. The syntax is straightforward - the body of function is defined 
directly in the `map*` function call. Try to rewrite the call of `map_dbl`
using anonymous function:
```{r map-lambda-example, exercise=TRUE, exercise.setup = "purrr-data-example"}
map_dbl(df, ___(x) {___(x)/___(x)})
```

```{r map-lambda-example-solution}
map_dbl(df, function(x) {sum(x)/length(x)})
```

```{r map-lambda-example-check}
lecture_grade_code()
```

### Shortcuts
In addition to the normal call of anonymous functions you can use "shortcut syntax"
for defining the function. The function body is defined using `~` operator (instead of `function(x)`). 
The first argument is passed to the function as `.`. The body of the function is defined after the `~` operator.
Rewrite the previ0us example using the shortcut syntax:
```{r map-lambda-example-2, exercise=TRUE, exercise.setup = "purrr-data-example"}
map_dbl(df, ___ sum(___)/length(___))
```

```{r map-lambda-example-2-solution}
map_dbl(df, ~ sum(.)/length(.))
```

```{r map-lambda-example-2-check}
lecture_grade_code()
```

## Data manipulation
Now we know a lot about the basic R and few things from the advanced topics such as
pipes. Finally, we can get to the modern approach to data analysis within the R using 
packages and functions from the universe of packages called `tidyverse`. It is the collection 
of packages designed for the data science. They share the same philosophy, grammar
and structures. For detailse [see](https://www.tidyverse.org/). The `tidyverse` package
is just the umbrella package, which will install all the packages usually 
needed for the data manipulation and analysis.

```{r tidyverse-setup, include=FALSE}
library(tidyverse)
```

### tibble
The `tibble` is core package of the whole `tidyverse`. It provides the `tibble` 
data structure, which is modern version of `data.frame`. It is more user friendly
and it is designed to be used with the `tidyverse` packages. You can create new `tibble`
using the `tibble` function. Try to create `tibble` with 3 columns `a`, `b`, `c`. Column
`a` should contain numbers from 1 to 4, column `b` should contain the number 3 and
column `c` should contain the sum of `a` squared and `b` columns. Do it now:

```{r tibble-example, exercise=TRUE}
tibble(a = ___,
       b = ___,
       c = ___)
```

```{r tibble-example-solution}
tibble(a = 1:4,
       b = 3,
       c = a^2 + b)
```

```{r tibble-example-check}
lecture_grade_code()
```

You can also convert existing `data.frame` to `tibble` using the `as_tibble` function. Let us
convert the `iris` data set to `tibble`:
```{r as_tibble, exercise=TRUE}
___(iris)
```

```{r as_tibble-solution}
as_tibble(iris)
```

```{r as_tibble-check}
lecture_grade_code()
```

The `tibble` will never change the column names and never creates row names (which is
somewhat strange behavior of `data.frame`). In addition, you can work with `tibble`
similarly to `data.frame`. 

### readr
To read data from files you can use base R function but there is `tidyverse` alternative, 
which is faster (approx. 10x); produce `tibble`s; will not produce factors from 
character columns; and does not depends on the OS and system variables (better reproducibility).
The most important functions from the package are:

- `read_csv` for reading the file
```{r read-iris, eval=FALSE}
data <- read_csv("iris.csv")
```
- `write_csv` for writing the file
```{r write-iris, eval=FALSE}
write_csv(iris, "iris.csv")
```

## Data wrangling
Data wrangling is the term, which represents reading, cleaning and transforming 
the data in order to be able to visualize and model them. In this section we 
will uncover the most common and useful function for manipulating tabular data (aka `data.frame`),
which are the most common structured data.

Data in appropriate format for analysis are called tidy. Tidy data set contains data
in tabular format where:

- each variable (feature) have own column
- each observation (sample) own row
- each value have own cell

`tidyverse` provides the set of functions, which are designed to work with tidy data.
In the following sections we will show the most important functions. Feature of the 
`tidyverse` functions is that you do not need to put names of columns in quotes because
of use the so-called `non-standard evaluation`. This means that the function will 
evaluate the column names and will use them in the function call.

### Iris dataset
The `iris` data set is the most famous data set in the world. It contains 150 samples
of iris flowers. Each sample is described by 4 features (sepal length, sepal width,
petal length, petal width) and the species of the flower (setosa, versicolor, virginica).
The data set is available in the base R. You can check the dataset bellow:

```{r iris}
iris
```

### mutate
`mutate` transforms the column values according to supplied formula and stores them
either to the original or to new column. In the first example adjust
the column `Petal.Length` to be squared: 
```{r mutate-example, exercise=TRUE}
iris |>
  ___(___ = ___)
```

```{r mutate-example-solution}
iris %>% 
  mutate(Petal.Length = Petal.Length^2)
```

```{r mutate-example-check}
lecture_grade_code()
```

Next, try to create new column `petal_area` containing the area of the petal calculated
as `Petal.Length * Petal.Width`:

```{r mutate-example-2, exercise=TRUE}
iris |>
  ___(___ = ___*___)
```

```{r mutate-example-2-solution}
iris |> 
  mutate(petal_area = Petal.Length*Petal.Width)
```

```{r mutate-example-2-check}
lecture_grade_code()
```

Finally, you can perform multiple "adjustments" in one `mutate` call. Try to 
achieve the same as in the previous two examples using only one `mutate` call:
```{r mutate-both-example, exercise=TRUE}
iris |> 
  mutate(___ = ___,
         ___ = ___)
```

```{r mutate-both-example-solution}
iris |> 
  mutate(Petal.Length = Petal.Length^2,
         petal_area = Petal.Length*Petal.Width)
```

```{r mutate-both-example-check}
lecture_grade_code()
```

Note, that in the last example the results of the first operation are used in the second
operation. This is the so-called "lazy evaluation". Thus the results in our second
example and the last one are different.

### select
`select` function is used to select the columns from the input `data.frame`. 
It is similar to the `select` statement in `SQL`. You can specify the columns
using their names or using the column numbers. The function will return a new `data.frame`
with the selected columns. Select the columns `Petal.Length` and `Species` from the `iris` data set:

```{r select-by-name-example, exercise=TRUE}
iris |>
  ___(___, ___)
```

```{r select-by-name-example-solution}
iris |> 
  select(Petal.Length, Species)
```

```{r select-by-name-example-check}
lecture_grade_code()
```

Another option to specify the columns is to use `:` operator. This operator
is used to specify the range of columns. Select the columns `Petal.Length` to `Species`
from the `iris` data set:
```{r select-range-example, exercise=TRUE}
iris |>
  ___(___:___)
```

```{r select-range-example-solution}
iris |> 
  select(Petal.Length:Species)
```

```{r select-range-example-check}
lecture_grade_code()
```

You can combine the `select` function with auxiliary functions, which help
you to select the columns based on some expression. The most important ones are: 
`starts_with`, `ends_with`, `contains`, `matches`, `everything`. Test the first auxiliary function
`starts_with` on the `iris` data set. Select all columns starting with `Petal`:
```{r select-starts-with-example, exercise=TRUE}
iris |>
  ___(___("___"))
```

```{r select-starts-with-example-solution}
iris |> 
  select(starts_with("Petal"))
```

```{r select-starts-with-example-check}
lecture_grade_code()
```

Sometimes is useful to deselect the columns. This can be done using `-` sign. Remove
column `Petal.Length` from the `iris` data set:
```{r select-deselect-example, exercise=TRUE}
iris |>
  ___(-___)
```

```{r select-deselect-example-solution}
iris |> 
  select(-Petal.Length)
```

```{r select-deselect-example-check}
lecture_grade_code()
```

The final use of the select function is to reorder the columns. This can be done
in combination with `everything` function. Move the `Species` column to the first position:
```{r select-reorder-example, exercise=TRUE}
iris |>
  ___(___, everything())
```

```{r select-reorder-example-solution}
iris |> 
  select(Species, everything())
```

```{r select-reorder-example-check}
lecture_grade_code()
```

### rename
`rename` function renames the column and it is in principle shortcut for `mutate` followed
by `select`. The use is similar to mutate function. Rename the column `Species` to `class`:
```{r rename-example, exercise=TRUE}
iris |>
  ___(class = ___)
```

```{r rename-example-solution}
iris |> 
  rename(class = Species)
```

```{r rename-example-check}
lecture_grade_code()
```

### filter
Very often you will need to filter the data set based on some condition. That 
means you want to select only several rows. The `filter` function is used
for this purpose. It is similar to `where` clause in `SQL`. Filter the `iris` data
set to keep only the rows with `Petal.Width` equal to `0.2`:

```{r filter-example, exercise=TRUE}
iris |>
  ___(___ == ___)
```

```{r filter-example-solution}
iris |> 
  filter(Petal.Width == 0.2)
```

```{r filter-example-check}
lecture_grade_code()
```

If you need to filter the data set based on multiple conditions you can use
`&` and `|` operators. In addition default is `&` operator. Thus you can  
order the conditions using `,` to chain them using and-logic. Try to filter
the `iris` data set to keep only the rows with `Petal.Width` equal to `0.2`
and `Sepal.Length` greater than `4.5`:

```{r filter-chain-example, exercise=TRUE}
iris |>
  ___(___ == ___, ___ > ___)
```

```{r filter-chain-example-solution}
iris |> 
  filter(Petal.Width == 0.2, Sepal.Length > 4.5)
```

```{r filter-chain-example-check}
lecture_grade_code()
```

### summarise
Computing summary statistics is one of the most common tasks in data analysis. To
achieve this you can use `summarise` function, which then computes the selected
statistics over the selected column(s). Try to extract the `max`imum of the `Petal.Length`:

```{r summarise-example, exercise=TRUE}
iris |>
  ___(max_petal_length = ___(___))
```

```{r summarise-example-solution}
iris |> 
  summarise(max_petal_length = max(Petal.Length))
```

```{r summarise-example-check}
lecture_grade_code()
```

### group_by
In the previous example we computed the maximum of the `Petal.Length` over the whole
data set. But we can also compute the maximum over the groups. The `group_by` function
is used to group the data set by the selected column(s). The `summarise` function
will then compute the statistics over the groups. The `group_by` function is similar
to the `group by` clause in `SQL`. Try to compute the maximum of the `Petal.Length`
for each `Species` in the `iris` data set:

```{r group_by-example, exercise=TRUE}
iris |>
  group_by(___) |>
  ___(max_petal_length = ___(___))
```

```{r group_by-example-solution}
iris |> 
  group_by(Species) |> 
  summarise(max_petal_length = max(Petal.Length))
```

```{r group_by-example-check}
lecture_grade_code()
```

## Reshaping data
One of the difficult tasks for example in `SQL` is reshaping the data from long to 
wide format (~ converting rows to columns and back). For that purpose there exists
two functions within the `tidyverse` - `pivot_wider` and `pivot_longer`.To
demonstrate their capabilities we will use "example" tables from the `tidyverse`. 

### pivot_longer

Check the table `table4a` to see where the problem is:
```{r table4a}
table4a
```

This table clearly contains two columns, which represents number of cases conditioned by variable - `year`. 
To make table tidy we need to transform it to have one column with the year and 
one column with the number of cases. Try it now:
```{r pivot-longer-example, exercise=TRUE}
table4a |>
  pivot_longer(___, # indexes of the columns to be transformed
               names_to = "___", # name of the new column with the names of the columns
               values_to = "___") # name of the new column with the values of the columns
```

```{r pivot-longer-example-solution}
table4a |>
  pivot_longer(2:3, # indexes or column names of the columns to be transformed
               names_to = "year", # name of the new column with the names of the columns
               values_to = "cases") # name of the new column with the values of the columns
```

```{r pivot-longer-example-check}
lecture_grade_code()
```


### pivot_wider

In our second table `table2` we have the opposite problem. The table is in long format
but it should be in wide format. The `type` column represents two columns with the values
from the `count` column. 
```{r table2}
table2
```

Use `pivot_wider` function to transform the table:
```{r pivot-wider-example, exercise=TRUE}
table2 |>
  pivot_wider(names_from = ___, # name of the column with the names of the columns
              values_from = ___) # name of the column with the values of the columns
```

```{r pivot-wider-example-solution}
table2 |>
  pivot_wider(names_from = type, # name of the column with the names of the columns
              values_from = count) # name of the column with the values of the columns
```

```{r pivot-wider-example-check}
lecture_grade_code()
```

## Joining data
Final part of the data manipulation is joining two tables together using the shared
key column. The `tidyverse` provides the set of functions for `join`ing the tables.

### Example data
We will use two example tables `x` and `y` created using `tribble` function, which
is part of the `tibble` package. It is used to create small tables in a readable
way. The tables are created in the following way:
```{r join-tables-init}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)

y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```

```{r join-tables}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)

y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```

### inner_join
The `inner_join` function is used to join two tables together using the shared key
column. The `inner_join` function will keep only the rows with matching keys in both
tables. The result will be a new table with the columns from both tables. Try to
join the tables `x` and `y` using the `key` column:

```{r inner-join-example, exercise=TRUE, exercise.setup = "join-tables"}
___ |>
  ___(___, by = "___")
```

```{r inner-join-example-solution}
x |>
  inner_join(y, by = "key")
```

```{r inner-join-example-check}
lecture_grade_code()
```

### left_join
The second join function is `left_join`. It is used to join two tables together
using the shared key column. The `left_join` function will keep all rows from the
left table and only the rows with matching keys from the right table. Join tables
`x` and `y` using the `key` column:

```{r left-join-example, exercise=TRUE, exercise.setup = "join-tables"}
___ |>
  ___(___, by = "___")
```

```{r left-join-example-solution}
x |>
  left_join(y, by = "key")
```

```{r left-join-example-check}
lecture_grade_code()
```

### full_join
The `full_join` function is used to join two tables together using the shared key
column. The `full_join` function will keep all rows from both tables. The result
will be a new table with the columns from both tables. The rows with matching keys
will be merged together. The rows with non-matching keys will be filled with `NA` values.
Join tables `x` and `y` using the `key` column:

```{r full-join-example, exercise=TRUE, exercise.setup = "join-tables"}
___ |>
  ___(___, by = "___")
```

```{r full-join-example-solution}
x |>
  full_join(y, by = "key")
```

```{r full-join-example-check}
lecture_grade_code()
```

### anti_join
The last join function we will be covering is `anti_join`. It provides you with 
the filtering functionality. It will keep only the rows from the left table, which
are not present in the right table. The result will be a new table with the columns
from the left table. Join tables `x` and `y` using the `key` column:

```{r anti-join-example, exercise=TRUE, exercise.setup = "join-tables"}
___ |>
  ___(___, by = "___")
```

```{r anti-join-example-solution}
x |>
  anti_join(y, by = "key")
```

```{r anti-join-example-check}
lecture_grade_code()
```

## Plotting
With the base R plots one can achieve many things, but the notation and the way how
the base plotting is handled is very old and little bit rusty. Lucky for us, there
exists package `ggplot2`, which provides the R with the so-called grammar of graphics
implementation. This grammar defines in rigorous way how the scientific figure should 
be build and how to achieve this with layering of the representation. In this section 
we will try to produce graphs we created in our previous introductory session.

```{r ggplot-setup, include=FALSE}
library(ggplot2)
```

### scatter plot
`ggplot2` works with so-called layers. Each layer defines the part of the figure. The
first layer is the data layer, which defines the data set and the aesthetics (x and y axis, etc.) of the
plot. The second layer is the geometry layer, which defines the type of the plot.

At first, we will create the simple scatter plot of the `iris` data set. Create
the scatter plot with `Petal.Length` on the x axis and `Petal.Width` on the y axis.
As the geometry we will use the `geom_point` function, which is used to create
the scatter plot. All functions from the `ggplot2` package are connected using `+` pipe operator.
Try to create the scatter plot:
```{r scatter-simple, exercise=TRUE}
ggplot(iris, # data
       # aeshetics
       aes(x = ___, 
           y = ___)) +
  ___() # geometry
```

```{r scatter-simple-solution}
ggplot(iris,
       aes(x = Petal.Length,
           y = Petal.Width)) +
  geom_point()
```

```{r scatter-simple-check}
lecture_grade_code()
```

By adding other layers we can extend the plot. Let us add color to the points
based on the `Species` column, include the regression line using `geom_smooth`,
and change the background theme to black and white using `theme_bw` function. 
Extend the plot now:
```{r scatter-extended, exercise=TRUE}
ggplot(iris,
       aes(x = ___,
           y = ___,
           color = ___)) +
  ___() + # scatter plot
  ___(method = "lm") + # regression line
  ___() # theme
```

```{r scatter-extended-solution}
ggplot(iris,
       aes(x = Petal.Length,
           y = Petal.Width,
           color = Species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

```{r scatter-extended-check}
lecture_grade_code()
```

### histogram
To produce the histogram similar (+/-) to the base R histogram, we need to just
adjust the code from the simple scatter plot. Remove the `y` aesthetics and change
the geometry to `geom_histogram`:

```{r histogram-simple, exercise=TRUE}
ggplot(iris,
       aes(x = Petal.Length,
           y = Petal.Width)) +
  geom_point()
```

```{r histogram-simple-solution}
ggplot(iris,
       aes(x = Petal.Length)) +
  geom_histogram()
```

```{r histogram-simple-check}
lecture_grade_code()
```

The histogram can be extended in the same way as the scatter plot. Add the `Species` column
to the `fill` aesthetics to color the histogram based on the species, add
the `position` parameter to the `geom_histogram` function to change the
position of the histogram to `identity` (un-stacked histogram) and add the `alpha` parameter to change
the opacity of the histogram. Finally use `theme_bw` to change the background.
Do it now:

```{r histogram-extended, exercise=TRUE}
ggplot(iris,
       aes(x = ___,
           fill = ___)) +
  ___(position = "___", ___ = 0.7) + # position and alpha
  ___() # theme
```

```{r histogram-extended-solution}
ggplot(iris,
       aes(x = Petal.Length,
           fill = Species)) +
  geom_histogram(position = "identity", alpha = 0.7) +
  theme_bw()
```

```{r histogram-extended-check}
lecture_grade_code()
```

## Conclusion
This concludes our second seminar lecture, where you learned about the advanced 
R topics such as pipes, `purrr` package and `tidyverse`. Next comes the second assignment,
which will available after of our seminar.

Here we repeat the important information regarding your consent to the data collection. The following hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:

**`r HASH`**

Please, keep it for your records in case you change your mind and want to delete your data.