---
title: "Data manipulation in R - assignemnt"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE, context="server"}
library(learnr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

library(readr)
library(tidyverse)
library(ggplot2)
```

# Assigment
In our second assignment we will be reworking the code from the previous assignment
using `tidyverse` packages. To recall - we will be processing subset the ASSISTments data set. 
And we will be creating two functions For us the data do not matter that much since we are going 
to create two functions: first will read the data from file, process them and
outputs the results in other file; and the second will read the original data 
and produce histogram of selected column. The instructions are the similar to
previous task :)

## I: read_transform_write

### Function 
Create function `read_transform_write`, which take one argument: `input`, which is a data.frame and returns it. Test the function call:
```{r exercise-setup, echo=FALSE}
input_data <- read_csv("data/assigment_data.csv")
```

```{r exercise-1, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___
}

read_transform_write(input_data)
```

### Select
`select` the columns `user_id`,`assignment_id`,`start_time`,`end_time` and store them in new `data.frame` named `result`.
Return the `result`:
```{r exercise-2, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___ <- 
    ___ |>
    ___(___)
  ___
}

read_transform_write(input_data)
```

### New columns
Create new column `elapsed_time` in `result` containing time difference between `end_time` and `start_time`, round the difference to whole number; and column `message` in `result` containing string: `"The user <user_id> finished assignment <assignment_id> in <elapsed_time> seconds.`
```{r exercise-3, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___ <- 
    ___ |>
    ___(___) |>
    ___(___ = ___(___(___,___)),
        ___ = ___(___))
  ___
}


read_transform_write(input_data)
```

Hints: 

  - it is possible to subtract two POSIXct values dt1-dt2 using `difftime` function, the result is in seconds
  - check function `round` 
  - for building the string you might search for the function `paste` or `paste0`
    
### Filter
The function should return only first 5 rows of the `result` data.frame. Adjust the code:
```{r exercise-4, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___ <- 
    ___ |>
    ___(___) |>
    ___(___ = ___(___(___,___)),
        ___ = ___(___)) |>
    ___(___ <= ___)
  ___
}

read_transform_write(input_data)
```
Hint: check function `row_number()`, which can be used within the `filter()`

## II: plot_histogram

### Function
Create function `plot_histogram`, which takes three arguments `df` data.frame, `col` string, `bins` number and returns `invisible(NULL)`:
```{r histogram-1, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___){
  ___
}

plot_histogram(input_data, "attempt_count", 3)
```

### Check column
Check that the `col` is a column in `df` and that it is numeric. If not, throw an error (you can test it with column names `attempt_count`, `type` and `non-existing-column`):
```{r histogram-2, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___) {
  if(!(___ %in% ___(___))){
    stop("Column does not exists")
  }
  if (!(___(___[[___]]))) {
    stop("Wrong column, values must be numeric")
  }
  ___
}

plot_histogram(input_data, "attempt_count", 3)
```
Hints: 
- you can throw error using `stop` function.
- to get names of the data.frame columns use `colnames` function

### Plot histogram
Now it is time to plot the histogram. Plot the histogram using the `geom_hist` of `col`umn and set the number of `bins`:
```{r histogram-3, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___) {
  if(!(___ %in% ___(___))){
    stop("Column does not exists")
  }
  if (!(___(___[[___]]))) {
    stop("Wrong column, values must be numeric")
  }
  ___ |> 
    ___(___(___ = ___)) +
    ___(___ = ___)
}

plot_histogram(input_data, "attempt_count", 3)
```
Hint: to ease ingestion of the string variable within the `aes`thetics you can 
use function `aes_string`.

### Pretty histogram
Finally, make the histogram pretty. Set the title of the histogram to `Histogram of column <col>`, set the color of the bars to `azure3` and change the `theme` of the histogram to `_bw`:
```{r histogram-4, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___) {
  if(!(___ %in% ___(___))){
    stop("Column does not exists")
  }
  if (!(___(___[[___]]))) {
    stop("Wrong column, values must be numeric")
  }
  ___ |> 
    ___(___(___ = ___)) +
    ___(___ = ___, ___ = "___") +
    ___(___("___",___)) +
    ___()
}

plot_histogram(input_data, "attempt_count", 3)
```
Hint: there exist function `ggtitle`.


## Submission 
**Now, you have the final result. Copy your code of `read_transform_write` and `plot_histogram` function and paste it 
to the Moodlelab in the [second assignment](https://moodlelab.hu-berlin.de/mod/assign/view.php?id=11432).**
