---
title: "Data manipulation in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---
```{r setup, include=FALSE, context="server"}
library(learnr)
library(gradethis)
library(glue)
library(dplyr)
library(curl)
library(jsonlite)
library(openssl)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

req <- curl_fetch_memory("https://api.ipify.org?format=json")

HASH <- as.character(md5(fromJSON(rawToChar(req$content))[[1]]))

event_recorder <- function(tutorial_id, 
                           tutorial_version,
                           user_id,
                           event, 
                           data) {
  REPO <- "C:/Users/annak/Documents/kuba_prace/learnr/"
  
  if (file.exists(glue("{REPO}/data_{HASH}.rds"))) {
    t <- readRDS(glue("{REPO}/data_{HASH}.rds"))
  } else {
    t <- 
      tibble(
        time = .POSIXct(numeric(0)),
        tutorial_id = character(),
        tutorial_version = character(),
        user_id = character(),
        hash = character(),
        event = character(),
        data = list()
      )
  }
  
  t <- bind_rows(t, 
                 tibble(
                   time = Sys.time(),
                   tutorial_id = tutorial_id,
                   tutorial_version = tutorial_version,
                   user_id = user_id,
                   hash = HASH,
                   event = event,
                   data = list(data)
                 )
  )
  
  saveRDS(t, file = glue("{REPO}/data_{HASH}.rds"))
}

options(tutorial.event_recorder = event_recorder)

lecture_grade_code <- function() {
  grade_this_code(
    correct = "Correct!",
    incorrect = "Try again. {code_feedback()}"
  )
}
```

## Introduction
At the end of this lecture you will:

- learn how to use pipes for improved code readability
- find out alternative to `apply` family 
- learn the basics of data wrangling
- explore the basics of advanced plotting in R
  
> For simplicity, we will not be using the R and RStudio installed on your computer. We will be
using interactive excercises created using [learnr](https://cran.r-project.org/web/packages/learnr/index.html) package.

> When you find `___` in the code chunk, it means that you need to fill in the missing part of the code.

## Informed Consent
This informed consent form is for a project investigating the interactions with the online
tutorial platform created using `learnr` package.

Name of Investigator: Dr. Jakub Kuzilek (jakub.kuzilek@hu-berlin.de)

Name of Supervisor: Prof. Dr. Niels Pinkwart (pinkwart@hu-berlin.de)

Name of Institution: Humboldt-Universität zu Berlin

**You can download this informed consent in the Moodle course.**

**The data from multiple tutorials will be collected and analysed.The informed consent is the same. For each session, you will get hash, which is the only unique identifier of your interactions with the material.** 

### Information sheet

#### Introduction of investigators
Jakub Kuzilek is a researcher in fields of educational data mining and learning analytics since completion of his PhD in Artificial Intelligence and Biocybernetics at the Czech Technical University. In 2020 he joined Prof. Dr. Pinkwarts's research group and currently working on the project IMPACT.

#### Purpose of the study
The purpose of this study is to investigate the interactions with the online tutorial platform created using the `learnr` package. The data will be used to improve the design of the tutorial and to understand how users interact with the platform.

#### Type of Research Intervention
To support and investigate these learning and interaction processes interaction data will be used. In addition performance data of the students will be collected.

#### Research Data Management
The collected data will be anonymized and merged. The collected data are stored on the servers of the Computer and Media Service (CMS) of the Humboldt-Universität zu Berlin in Germany and the computers of the researchers. The data access is protected and regulated via a login and the access is restricted to the members of the project team. The personal data will be stored on secured location accessible via authenticated encrypted connection (VPN and SSH). The data are archived for another 10 years after the end of the project according to good scientific practice. The anonymized data may be published to enable Open Access for the general public. The participant may gain access to his/her recorded data, request a correction to his/her personal data record, or withdraw from the research study by contacting the Investigator:
Jakub Kuzilek at jakub.kuzilek@hu-berlin.de

In case of doubt as to the lawfulness of the processing of personal data, data subjects have the right to lodge a complaint with a Member State supervisory authority. The competent supervisory authority for Humboldt-Universität zu Berlin is Die Berliner Beauftragte für Datenschutz und Informationsfreiheit, Friedrichstraße 219, 10969 Berlin. The official data protection officer for Humboldt-Universität zu Berlin is: Gesine Hoffmann-Holland, Tel: +49 (30) 2093-20022, datenschutz@uv.hu-berlin.de.

#### Voluntary participation
Participation in this research project is entirely voluntary. It is up to the participant to decide whether or not to participate and there is no explicit or implicit obligation to participate. The participant has the possibility to cancel the participation at any time - even after the signing of the Informed Consent - without any disadvantages for the participant. 

#### Documentation of the consent
One copy of this document will be kept separately from the research records of this study using a pseudonym id in order to remove data after the request from the participant. The contact details of the participants will be deleted after the project is finished. To assure data deletion of the requesting subject a unique identifier in form of the email address and is required.

### Certificate of consent
- I agree to the collection of my pseudonymized interaction data in the course "Machine Learning for Education".
- I agree to the collection of my pseudonymized perfomance data in the course "Machine Learning for Education".
- I agree to have the anonymized interaction data and the anonymized performance data published under the Creative Commons Attribution 4.0 International (CC-BY) Licence (https://creativecommons.org/licenses/ by/4.0/) and so I allow it to be used by other people for other projects. 
- I agree to have the interaction data, and the performance data archived for 10 years after finishing of the project.

This hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:

**`r HASH`**

Please, keep it for your records in case you change your mind and want to delete your data.

```{r informed-consent, echo=FALSE}
question_radio("I agree to the collection of my pseudonymized interaction data in the course 'Machine Learning for Education'.", 
               correct = NULL, 
               answer("yes", correct = TRUE),
               answer("no", correct = TRUE)
)
```


## Advanced R programming
We will introduce two topics, which can be considered as advanced, but 
at the end both approaches simplify the programming in R. The first topic are 
`pipes`, which works in similar way to the pipes in the Linux bash command line. 
They enable chaining commands (functions) in more readable format with only a little
overhead compared to the normal function call. The second topic is package
`purrr`, which is the alternative to the apply function family providing more consistent
way for handling the iteration using functional programming paradigm.

## Pipes

Pipes are the way how to chain the function calls together. The main idea is to
pass the result of one function to the next function as an argument. There are two
types of pipes in R. The first one is the native pipe, which is available in R
version 4.1.0 and higher. The second one is the pipe operator `%>%`, which
comes from the `magrittr` package. It is also available in the `dplyr` package, 
which is part of the `tidyverse`. 

### Example 
Let us start with the example. Imagine you are writing the program for cooking pasta, 
each step is represented by one function. The input to the first function `get_pasta`
would be `type` of pasta and `number_of_people`, who are going to eat that pasta. The output
is a list containing the type of pasta and number of people. Try to define the function:

```{r pipes-get-pasta-example, exercise=TRUE}
___ <- function(___, ___) {
  print(paste0("Getting ",___," for ",___," persons."))
  list(___, ___)
}
```

```{r pipes-get-pasta-example-solution}
get_pasta <- function(type, number_of_people) {
  print(paste0("Getting ",type," for ",number_of_people," persons."))
  list(type, number_of_people)
}
```

```{r pipes-get-pasta-example-check}
lecture_grade_code()
```

Now we can define functions `cook_pasta` and `serve_pasta`. The first one will take the
pasta list and print the message about cooking. It will also `append` the cooking time
to the list and returns it. The second one will take the pasta list and print the message about serving
the pasta. It will return the same list as an output. Try to define the functions:
```{r pipes-cook-serve-pasta-example, exercise=TRUE}
___ <- function(___) {
  print(paste0("Cooking ", ___[[1]],".")) # get the type of pasta (1st element of the list)
  ___(___, "10 minutes")
}

___ <- function(___) {
  # get the type of pasta (1st element of the list),
  # number of people (2nd element of the list), 
  # cooking time (3rd element of the list)
  print(paste0("Serving ", ___[[___]], " for ", ___[[___]], "persons cooked for ", ___[[___]],"."))
  ___
}
```

```{r pipes-cook-serve-pasta-example-solution}
cook_pasta <- function(pasta) {
  print(paste0("Cooking ", pasta[[1]], "."))
  pasta[[3]] <- "10 minutes"
  pasta
}

serve_pasta <- function(pasta) {
  # get the type of pasta (1st element of the list),
  # number of people (2nd element of the list), 
  # cooking time (3rd element of the list)
  print(paste0("Serving ", pasta[[1]], " for ", pasta[[2]], " persons cooked for ", pasta[[3]], "."))
  pasta
}
```

```{r pipes-cook-serve-pasta-example-check}
lecture_grade_code()
```

### Cooking using variables
Normally, we will call the functions in defined order and store the interim results
in variables. Those variables can be used later in the code. Try this in the following
exercise. Get the `spaghetti` pasta for 20 people, cook it and serve it:
```{r pipes-pasta}
get_pasta <- function(type, number_of_people) {
  print(paste0("Getting ",type," for ",number_of_people," persons."))
  list(type, number_of_people)
}

cook_pasta <- function(pasta) {
  print(paste0("Cooking ", pasta[[1]], "."))
  pasta[[3]] <- "10 minutes"
  pasta
}

serve_pasta <- function(pasta) {
  # get the type of pasta (1st element of the list),
  # number of people (2nd element of the list), 
  # cooking time (3rd element of the list)
  print(paste0("Serving ", pasta[[1]], " for ", pasta[[2]], " persons cooked for ", pasta[[3]], "."))
  pasta
}
```

```{r pipes-objects-example, exercise=TRUE, exercise.setup = "pipes-pasta"}
pasta <- ___("___", ___)
cooked_pasta <- ___(pasta)
result <- ___(cooked_pasta)
```

```{r pipes-objects-example-solution}
pasta <- get_pasta("spaghetti", 20)
cooked_pasta <- cook_pasta(pasta)
result <- serve_pasta(cooked_pasta)
```

```{r pipes-objects-example-check}
lecture_grade_code()
```

### Cooking compressed
We can call function from function to pass the interim results as an argument. This 
will save some space and typing. Try to do the same as in the previous exercise:
```{r pipes-compressed-example, exercise=TRUE, exercise.setup = "pipes-pasta"}
result <- ___(___(___("___", 20)))
```

```{r pipes-compressed-example-solution}
result <- serve_pasta(cook_pasta(get_pasta("spaghetti", 20)))
```

```{r pipes-compressed-example-check}
lecture_grade_code()
```

### Cooking using pipes
The previous two examples are the standard way how to call functions in R. Both ways
are correct and they are used in the R code. But they are not very readable and
are sensitive for typing errors and somewhat are hard to read. Especially the 
second case. For hard readable code we got Lisp :) Luckily, we can use the pipes
to help us. We will be using base R pipe `|>` operator in our examples. Try to 
rewrite the previous example using the pipe:

```{r pipes-base-pipe-example, exercise=TRUE, exercise.setup = "pipes-pasta"}
result <-
  ___("spaghetti", 20) |> 
  ___() ___
  ___()
```

```{r pipes-base-pipe-example-solution}
result <-
  get_pasta("spaghetti", 20) |>
  cook_pasta() |>
  serve_pasta()
```

```{r pipes-base-pipe-example-check}
lecture_grade_code()
```

This form of writing the code helps you focus on verbs (=function names) and you
can read the function composition as a set of imperative actions. Not all the times 
using a pipe is useful, but in the most cases it improves the code readability and 
helps you with the sequences of linear operations. 

## Advanced iterations

The `purrr` package comes as an alternative to the `apply` function family. In 
most cases it provides the same functionality and performance like `apply` functions,
but it comes with more consistent function calls and names, thus it is easier to 
understand and work with. 

```{r purrr-setup, include=FALSE}
library(purrr)
```

### Example data
Before we start using the `map*` functions from `purrr`. We need to create example `data.frame` `df`
with 4 columns (`a`,`b`,`c`,`d`) and `10` rows. The columns will be filled with random numbers from normal
distribution using `rnorm` function. Create the `df` data.frame:
```{r purrr-data-example, exercise=TRUE}
___ <- ___(
  ___ = rnorm(___),
  ___,
  ___,
  ___
)
```

```{r purrr-data-example-solution}
df <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

```{r purrr-data-example-check}
lecture_grade_code()
```

### Apply refresher
You are familiar with the `apply` function family from the previous lecture. Use it
on the `df` `data.frame` to calculate the `mean` of each column (`MARGIN=2`, second argument):
```{r apply-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
___(___,2,___)
```

```{r apply-example-solution}
apply(df, 2, mean)
```

```{r apply-example-check}
lecture_grade_code()
```

### map
The `map` family of functions is the alternative to the `apply` function family. To call
the base function `map` you need to provide the first argument, which is the input
data (e.g. `data.frame`), the second argument is the function you want to apply.
Try to rewrite the previous example:
```{r map-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
___(___, ___)
```

```{r map-example-solution}
map(df, mean)
```

```{r map-example-check}
lecture_grade_code()
```

### map family
The `map` function is the most general one. It can be used with any type of input
data. It returns the result as a list. If one wants to have the result in a specific
format, the only thing you need to do is to add the suffix to the function name. 
In our case we want to have the result in a numeric vector format thus we can
add `_dbl` suffix to the function name. Try to modify previous example:
```{r map-dbl-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
map(df, mean)
```

```{r map-dbl-example-solution}
map_dbl(df, mean)
```

```{r map-dbl-example-check}
lecture_grade_code()
```

From the function name we know the output and the call is consistent
with other functions from the "family".The other fuctions from family are: 
`map_lgl`, `map_int`, `map_chr`.

### Own functions
Both `apply` and `map` family can work with your own defined functions. Define your
own function `my_mean` using `sum` and `length` functions and use it in the call
of `map_dbl` function:
```{r map-own-function-example, exercise=TRUE, exercise.setup = "purrr-data-example-solution"}
___ <- function(x) {
  ___(x)/___(x)
}

map_dbl(df, ___)
```

```{r map-own-function-example-solution}
my_mean <- function(x) {
  sum(x)/length(x)
}

map_dbl(df, my_mean)
```

```{r map-own-function-example-check}
lecture_grade_code()
```

### Anonymous functions
Within the call of the `map` function you can also use the anonymous functions. This 
is handy when the function is used only once and you do not want to define it
explicitly. The syntax is straightforward - the body of function is defined 
directly in the `map*` function call. Try to rewrite the call of `map_dbl`
using anonymous function:
```{r map-lambda-example, exercise=TRUE, exercise.setup = "purrr-data-example"}
map_dbl(df, ___(x) {___(x)/___(x)})
```

```{r map-lambda-example-solution}
map_dbl(df, function(x) {sum(x)/length(x)})
```

```{r map-lambda-example-check}
lecture_grade_code()
```

### Shortcuts
In addition to the normal call of anonymous functions you can use "shortcut syntax"
for defining the function. The function body is defined using `~` operator (instead of `function(x)`). 
The first argument is passed to the function as `.`. The body of the function is defined after the `~` operator.
Rewrite the previ0us example using the shortcut syntax:
```{r map-lambda-example-2, exercise=TRUE, exercise.setup = "purrr-data-example"}
map_dbl(df, ___ sum(___)/length(___))
```

```{r map-lambda-example-2-solution}
map_dbl(df, ~ sum(.)/length(.))
```

```{r map-lambda-example-2-check}
lecture_grade_code()
```

## Data manipulation
Now we know a lot about the basic R and few things from the advanced topics such as
pipes. Finally, we can get to the modern approach to data analysis within the R using 
packages and functions from the universe of packages called `tidyverse`. It is the collection 
of packages designed for the data science. They share the same philosophy, grammar
and structures. For detailse [see](https://www.tidyverse.org/). The `tidyverse` package
is just the umbrella package, which will install all the packages usually 
needed for the data manipulation and analysis.

```{r tidyverse-setup, include=FALSE}
library(tidyverse)
```

### tibble
The `tibble` is core package of the whole `tidyverse`. It provides the `tibble` 
data structure, which is modern version of `data.frame`. It is more user friendly
and it is designed to be used with the `tidyverse` packages. You can create new `tibble`
using the `tibble` function. Try to create `tibble` with 3 columns `a`, `b`, `c`. Column
`a` should contain numbers from 1 to 4, column `b` should contain the number 3 and
column `c` should contain the sum of `a` squared and `b` columns. Do it now:

```{r tibble-example, exercise=TRUE}
tibble(a = ___,
       b = ___,
       c = ___)
```

```{r tibble-example-solution}
tibble(a = 1:4,
       b = 3,
       c = a^2 + b)
```

```{r tibble-example-check}
lecture_grade_code()
```

You can also convert existing `data.frame` to `tibble` using the `as_tibble` function. Let us
convert the `iris` data set to `tibble`:
```{r as_tibble, exercise=TRUE}
___(iris)
```

```{r as_tibble-solution}
as_tibble(iris)
```

```{r as_tibble-check}
lecture_grade_code()
```

The `tibble` will never change the column names and never creates row names (which is
somewhat strange behavior of `data.frame`). In addition, you can work with `tibble`
similarly to `data.frame`. 

### readr
To read data from files you can use base R function but there is `tidyverse` alternative, 
which is faster (approx. 10x); produce `tibble`s; will not produce factors from 
character columns; and does not depends on the OS and system variables (better reproducibility).
The most important functions from the package are:

- `read_csv` for reading the file
```{r read-iris, eval=FALSE}
data <- read_csv("iris.csv")
```
- `write_csv` for writing the file
```{r write-iris, eval=FALSE}
write_csv(iris, "iris.csv")
```

## Data wrangling
Data wrangling is the term, which represents reading, cleaning and transforming 
the data in order to be able to visualize and model them. In this section we 
will uncover the most common and useful function for manipulating tabular data (aka `data.frame`),
which are the most common structured data.

Data in appropriate format for analysis are called tidy. Tidy data set contains data
in tabular format where:

- each variable (feature) have own column
- each observation (sample) own row
- each value have own cell

### mutate
`mutate` transforms the column values according to supplied formula and stores them
either to the original or to new column. Let us show the examples on `iris` data set:
```{r mutate}
iris %>% 
  mutate(Petal.Length = Petal.Length^2)

iris %>% 
  mutate(petal_area = Petal.Length*Petal.Width)
```

We can also do multiple operations at once:
```{r mutate-2}
iris %>% 
  mutate(Petal.Length = Petal.Length^2,
         petal_area = Petal.Length*Petal.Width)
```

### select
`select` selects the specified columns. You can also use it for column reordering. 
And there exist multiple helper functions, which simplify the process. Let us do 
some selection on the `iris` data set:
```{r select}
iris %>% 
  select(Petal.Length, Species)
```

We can also decide that we what to drop some columns using `-` sign:
```{r select-2}
iris %>% 
  select(-Petal.Length)
```

Finally, we would like to move Species column to be the first not last:
```{r select-3}
iris %>% 
  select(Species, everything())
```

> Try to explore what `everything` is doing and what are the other helper functions.

### rename
`rename` function renames the column and it is in principle shortcut for `mutate` followed
by `select`. Let us show the example:
```{r rename}
iris %>% 
  rename(class = Species)
```

> Try to achieve same renaming with the combination of `mutate` and `select` functions.

### filter
`filter` function subset the rows of the input `data.frame` it works similarly to 
`where` clause in `SQL`. Let us try it: 
```{r filter}
iris %>% 
  filter(Petal.Width == 0.2)
```

You can chain the conditions (which is equivalent to & operator):
```{r filter-chain}
iris %>% 
  filter(Petal.Width == 0.2, Sepal.Length > 4.5)
```

> Try to filter out all setosa iris flowers with petal length less or equal to 1.4.

### summarise
For summary operation over the selected column(s) you can use `summarise` function. 
For example:
```{r summarise}
iris %>% 
  summarise(max_petal_length = max(Petal.Length))
```

### group_by
If we need to take into account the categorical variables defining groups (aka class) in 
our data we can use `group_by` function. For example:
```{r group_by}
iris %>% 
  group_by(Species) %>% 
  summarise(max_petal_length = max(Petal.Length))
```

### Reshaping tables
One of the difficult tasks for example in `SQL` is reshaping the data from long to 
wide format (~ converting rows to columns and back). For that purpose there exists
two functions within the `tidyverse` - `pivot_wider` and `pivot_longer`. To
demonstrate their capabilities we will use "example" tables from the `tidyverse`. The first:
```{r table4a}
table4a
```
This table clearly contains two columns, which represenst one variable - `year`. Thus
it is viable to transform the table to include this variable using `pivot_longer`:
```{r pivot-longer}
table4a %>% 
  pivot_longer(2:3, names_to = "year", values_to = "cases")
```

And the second example table:
```{r table2}
table2
```

Again this table has problem, but it is opposite to the previous one - the column
`type` clearly represent two columns with values from column `count`. So we can transform
the table using `pivot_wider`:
```{r pivot-wider}
table2 %>% 
  pivot_wider(names_from = type, values_from = count)
```

### Joining tables
Sometimes you will need to join two `data.frame`s together. This usually involves
some key column, which is shared by both data sets. Let us create two sample `tibbles` first:
```{r join-tables-init}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)

y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```

Note the way how we created them. We used function `tribble`, which is shortcut
to transposed `tibble`. It makes the manual data entry task for small data more 
readable. 

Now we can perform some joins over both tables.

### inner_join
The `inner_join` joins the observations from x to y keeping only those with matching
`key`:
```{r inner-join}
x %>% 
  inner_join(y, by = "key")
```

> Try to join the tables withou specifying the `by` argument.
> There exist function `semi_join` what it does? What is the difference between semi and inner joins?

### left_join
To match all observations from y to observations from x and keeping all x observations 
one can use `left_join`:
```{r left-join}
x %>% 
  left_join(y, by = "key")
```

> Is there opposite direction join?

### full_join
To merge both tables and keep all records with or without matching you can use
the `full_join`:
```{r full-join}
x %>% 
  full_join(y, by = "key")
```

### anti_join
Very useful "trick" for finding non-matching samples from one data set is provided 
by `anti_join`:
```{r anti-join}
x %>% 
  anti_join(y, by = "key")
```

# Plotting
With the base R plots one can achive many things, but the notation and the way how
the base plotting is handled is very old and little bit rusty. Lucky for us, there
exists package `ggplot2`, which provides the R with the so-called grammar of graphics
implementation. This grammar defines in rigorous way how the scientific figure should 
be build and how to achieve this with layering of the representation. 

We will try to produce graphs we created in our last introductory session. And on them
we will explain basics of plotting using `ggplot2`.

## Scatter plot
The first figure we have been producing is scatter plot of the `iris` data set. Let us
try to produce the original simple figure:
```{r scatter-simple}
ggplot(iris,
       aes(x = Petal.Length,
           y = Petal.Width)) +
  geom_point()
```

The first function `ggplot` defines the data layer of our plot. First argument informs
the plotting function about the data and the second argument is "function", which 
define the plot aesthetics. In this simple case we have two aesthetics defining 
data on x and y axis. This foundation data layer is then enhanced with the geometry
layer, telling the plotting function what kind of graph we are interested in. In
this case we would like to plot points, thus `geom_point`.

Finally, please note the `+` sign. It is `ggplot2` specific type of pipe, which cannot
be replaced with the ` %>% ` pipe.

We can extend this basic graph with more layers or adjustments. Let us do few more 
things:
```{r scatter-extended}
ggplot(iris,
       aes(x = Petal.Length,
           y = Petal.Width,
           color = Species)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

We added one more aesthetics color, which tells the plot how to determine colors
of the points. Also we, added one more geom_ layer, which plotted the linear regression
fit of each iris species. And finally we changed the background theme. 

## Histogram
To produce the histogram similar (+/-) to the base R histogram, we need to just
adjust the code from the previous plot:
```{r histogram}
ggplot(iris,
       aes(x = Petal.Length)) +
  geom_histogram() +
  theme_bw()
```

We are interested in one variable, thus we need just one aesthetics, which is 
the `x` and to plot the histogram one needs to add histogram geometry layer instead
of the point geometry. Again, we can extend the histogram using the similar logic
as in scatter plot:
```{r histogram-extended}
ggplot(iris,
       aes(x = Petal.Length,
           fill = Species)) +
  geom_histogram(position = "identity", alpha = 0.7) +
  theme_bw()
```
We added some additional parameters to the histogram: position forces the plotting
to produced "un-stacked" histogram and the alpha defines the opacity of the plotted
objects.

> Try to remove `position="identity"` from the function call. What happened?
> Change `fill` argument in `aes` to `color`. What happened?

## Conclusion
The base plotting concludes our introduction to R. Next comes the first assignment,
which will available after of our seminar.

Here we repeat the important information regarding your consent to the data collection. The following hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:
**`r HASH`**

Please, keep it for your records in case you change your mind and want to delete your data.

# Assigment
In our second assignment we will be reworking the code from the previous assignment
using `tidyverse` packages. To recall - we will be processing subset the ASSISTments data set. 
And we will be creating two functions For us the data do not matter that much since we are going 
to create two functions: first will read the data from file, process them and
outputs the results in other file; and the second will read the original data 
and produce histogram of selected column. The instructions are the similar to
previous task :)

## Instructions
0. Create new Rmd document called `02_submission.Rmd` and save it in the same folder as this file. Then
use this document to write the code for the assignment (use code chunks and comment them properly).
1. Create function `read_transform_write` which:
  - takes two arguments: `input_path` and `output_path` strings
  - assumes that the input is csv file in required format (you do not need to check that)
  - reads the file in the `input_path`
  - selects the columns `user_id`,`assignment_id`,`start_time`,`end_time` and store them in new `data.frame` named `result`
  - creates new column `elapsed_time` in `result` containing time difference between `end_time` and `start_time`, round the difference to whole number.
  - creates new column `message` in `result` containing string: `"The user <user_id> finished assignment <assignment_id> in <elapsed_time> seconds.`
  - stores the `result` to the csv file in `output_path`
  - prints out first 5 rows of the `result` 
  - returns 0 when finished 
  - Hints: 
    - to delete column from `data.frame` you can assign `NULL` value to the corresponding column: `df$x <- NULL`
    - it is possible to subtract two POSIXct values dt1-dt2, the result is in seconds
    - check function `round` 
    - for building the string you might search for the function `paste` or `paste0`
    - there exists functions for examining structure of `data.frame` such as `str`, `head`, `tail`
2. Call the function `read_transform_write` with the arguments `assigment_data.csv` and `task_1.csv"`. Please, note
that path might need to be changed based on the location of files and your system.
3. Create function `plot_histogram` which:
  - takes three arguments `df` data.frame, `col` string, `bins` number
  - extracts the data from `df` column `col` and plots the histogram with corresponding number of `bins`
    - histogram should have the title: `Histogram of column <col>`
    - the bars should be `azure3` color
    - the legend for the x axis should be the name of the column `col`
  - make sure that the histogram function is working only for columns where it makes sense, else 
  it should throw an error 
  - Hints:
    - to throw an error use function `stop`
4. Call function `plot_histogram` with data.frame `df` loaded from `assigment_data.csv`
and `col` set to `attempt_count` with `bins` set to `3`. 
    
## Submission 
Submit the task via Moodle in corresponding section. Your submission should contain
one `.Rmd` file named `02_submission.Rmd`. 