---
title: "Unsupervised Learning"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/custom.css
runtime: shiny_prerendered
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(glue)
library(dplyr)
library(curl)
library(jsonlite)
library(openssl)
library(tidyverse)
library(factoextra)
library(dendextend)
library(cluster)
library(patchwork)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

options(user_id = NULL)

lecture_grade_code <- function() {
  grade_this_code(
    correct = "Correct!",
    incorrect = "Try again. {code_feedback()}"
  )
}
```

```{js, context = "head", echo = FALSE}
// JavaScript code to generate a unique ID and send it to R
$(document).ready(function() {
  // Function to send the unique ID to R
  function sendIDToR(uniqueID) {
    // Store in localStorage for persistence
    localStorage.setItem('learnr_user_id', uniqueID);
    
    // Send to R via Shiny if available
    if (typeof Shiny !== 'undefined' && typeof Shiny.setInputValue === 'function') {
      Shiny.setInputValue('user_unique_id', uniqueID);
      console.log('IP address sent to R:', uniqueID);
    } else {
      // If Shiny isn't ready yet, try again in 500ms
      setTimeout(function() { sendIDToR(uniqueID); }, 500);
    }
  }
  
  // Check if we already have the ID in the localStorage
  const uniqueID = localStorage.getItem('learnr_user_id');
  if (uniqueID) {
    // If we have it, send it to R
    sendIDToR(uniqueID);
  } else {
    // Generate a new unique ID
    sendIDToR(Math.random().toString(36).substring(2, 15));
  }
});
```

```{r, context="server", echo=FALSE}
# IMPORTANT: This chunk runs in the Shiny server context
# Here 'input' object is available

# Create a reactive value to store the user's IP address
user_id <- reactiveVal(NULL)

# Observer that captures the IP address from JavaScript
observeEvent(input$user_unique_id, {
  # Store the IP address in our reactive value
  user_id(md5(input$user_unique_id))
  
  # Also store in options for global access
  id_data <- getOption("user_id")
  if (is.null(id_data)) {
    id_data <- md5(input$user_unique_id)
    options(user_id = id_data)
  }
})

# For debugging: Display the captured IP address
output$id_display <- renderText({
  if(is.null(user_id())) {
    "ID not yet generated"
  } else {
    user_id()
  }
})

output$id_display2 <- renderText({
  if(is.null(user_id())) {
    "ID not yet generated"
  } else {
    user_id()
  }
})

Sys.sleep(0.7)

event_recorder <- function(tutorial_id, 
                           tutorial_version,
                           user_id,
                           event, 
                           data) {
  REPO <- "/Users/jakubkuzilek/Documents/projects/temp"
  
  HASH <- getOption("user_id")
  
  if (file.exists(glue("{REPO}/data_{HASH}.rds"))) {
    t <- readRDS(glue("{REPO}/data_{HASH}.rds"))
  } else {
    t <- 
      tibble(
        time = .POSIXct(numeric(0)),
        tutorial_id = character(),
        tutorial_version = character(),
        user_id = HASH,
        hash = character(),
        event = character(),
        data = list()
      )
  }
  
  t <- bind_rows(t, 
                 tibble(
                   time = Sys.time(),
                   tutorial_id = tutorial_id,
                   tutorial_version = tutorial_version,
                   user_id = HASH,
                   event = event,
                   data = list(data)
                 )
  )
  
  saveRDS(t, file = glue("{REPO}/data_{HASH}.rds"))
}

options(tutorial.event_recorder = event_recorder)
```

## Introduction
At the end of this lecture you will:

- know how to perform clustering using  k-means algorithm (`kmeans`).
- select the number of clusters `k` for the k-means algorithm
- perform clustering using agglomerative hierarchical clustering (`hclust`)
- understand the structure of dendogram and the different distance computation
approaches

> For simplicity, we will not be using the R and RStudio installed on your computer. We will be
using interactive excercises created using [learnr](https://cran.r-project.org/web/packages/learnr/index.html) package.

> When you find `___` in the code chunk, it means that you need to fill in the missing part of the code.

## Informed Consent
This informed consent form is for a project investigating the interactions with the online
tutorial platform created using `learnr` package.

Name of Investigator: Dr. Jakub Kuzilek (jakub.kuzilek@hu-berlin.de)

Name of Supervisor: Prof. Dr. Niels Pinkwart (pinkwart@hu-berlin.de)

Name of Institution: Humboldt-Universität zu Berlin

**You can download this informed consent in the Moodle course.**

**The data from multiple tutorials will be collected and analysed.The informed consent is the same. For each session, you will get hash, which is the only unique identifier of your interactions with the material.** 

### Information sheet

#### Introduction of investigators
Jakub Kuzilek is a researcher in fields of educational data mining and learning analytics since completion of his PhD in Artificial Intelligence and Biocybernetics at the Czech Technical University. In 2020 he joined Prof. Dr. Pinkwarts's research group and currently working on the project IMPACT.

#### Purpose of the study
The purpose of this study is to investigate the interactions with the online tutorial platform created using the `learnr` package. The data will be used to improve the design of the tutorial and to understand how users interact with the platform.

#### Type of Research Intervention
To support and investigate these learning and interaction processes interaction data will be used. In addition performance data of the students will be collected.

#### Research Data Management
The collected data will be anonymized and merged. The collected data are stored on the servers of the Computer and Media Service (CMS) of the Humboldt-Universität zu Berlin in Germany and the computers of the researchers. The data access is protected and regulated via a login and the access is restricted to the members of the project team. The personal data will be stored on secured location accessible via authenticated encrypted connection (VPN and SSH). The data are archived for another 10 years after the end of the project according to good scientific practice. The anonymized data may be published to enable Open Access for the general public. The participant may gain access to his/her recorded data, request a correction to his/her personal data record, or withdraw from the research study by contacting the Investigator:
Jakub Kuzilek at jakub.kuzilek@hu-berlin.de

In case of doubt as to the lawfulness of the processing of personal data, data subjects have the right to lodge a complaint with a Member State supervisory authority. The competent supervisory authority for Humboldt-Universität zu Berlin is Die Berliner Beauftragte für Datenschutz und Informationsfreiheit, Friedrichstraße 219, 10969 Berlin. The official data protection officer for Humboldt-Universität zu Berlin is: Gesine Hoffmann-Holland, Tel: +49 (30) 2093-20022, datenschutz@uv.hu-berlin.de.

#### Voluntary participation
Participation in this research project is entirely voluntary. It is up to the participant to decide whether or not to participate and there is no explicit or implicit obligation to participate. The participant has the possibility to cancel the participation at any time - even after the signing of the Informed Consent - without any disadvantages for the participant. 

#### Documentation of the consent
One copy of this document will be kept separately from the research records of this study using a pseudonym id in order to remove data after the request from the participant. The contact details of the participants will be deleted after the project is finished. To assure data deletion of the requesting subject a unique identifier in form of the email address and is required.

### Certificate of consent
- I agree to the collection of my pseudonymized interaction data in the course "Machine Learning for Education".
- I agree to the collection of my pseudonymized perfomance data in the course "Machine Learning for Education".
- I agree to have the anonymized interaction data and the anonymized performance data published under the Creative Commons Attribution 4.0 International (CC-BY) Licence (https://creativecommons.org/licenses/ by/4.0/) and so I allow it to be used by other people for other projects. 
- I agree to have the interaction data, and the performance data archived for 10 years after finishing of the project.

This hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:

```{r, echo=FALSE}
textOutput("id_display")
```
\
Please, keep it for your records in case you change your mind and want to delete your data.

```{r informed-consent, echo=FALSE}
question_radio("I agree to the collection of my pseudonymized interaction data in the course 'Machine Learning for Education', lecture: 'Unsupervised Learning'.", 
               correct = NULL, 
               answer("yes", correct = TRUE),
               answer("no", correct = TRUE)
)
```

## Reproducibility
The clustering algorithms, especially k-means, are sensitive to the initial setup. Thus
you will never get the same result because of the random number generator embeded in R.
To fix the "seed" (random number generation start) you need to `set.seed` function. Set
the seed to the `1234`:
```{r set-seed, exercise=TRUE}
set.seed(___)
```

```{r set-seed-solution}
set.seed(1234)
```

```{r set-seed-check}
lecture_grade_code()
```

To check random number generator state you cang access the value via `.Random.seed`. Try 
to change seed to different number and check how the state changes.

We can check the random number generator state using the value in `.Random.seed`. Access
the state of the generator now:

```{r random-seed, exercise=TRUE, exercise.setup = "set-seed-solution"}
.Random.seed
```

In case you want to learn more about how the random number generator works in R, you can
check the [R documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Random.html).

## Data
For our lecture we will be using data set `USArrests` containing statistics in arrests per
100.000 residents in US in 1973. The data contains information about Murder, Assault
Rape and percentage of urban population living in urban areas. The data set is available
in R by default as `data.frame`. The name of the US state is not the data column in the 
`data.frame` but it is stored as row name. This is not recommended way how to work with data 
and it should be avoided. This data "issue" is due to legacy reasons. 

### `str`ucture
Now, we can explore the data. Start with the `str` function, which explores the 
structure of the `data.frame`:

```{r usarrests-str, exercise=TRUE}
___(___)
```

```{r usarrests-str-solution}
str(USArrests)
```

```{r  usarrests-str-check}
lecture_grade_code()
```

You can observe that each column of the `data.frame` is listed in the answer together
with the value types and first few values as an example. Finally the number of
observations and total number of variables (features/columns) is listed.

### `head`
To check the first few samples from the data, you can use function `head`. Try that 
now:

```{r usarrests-head, exercise=TRUE}
___(___)
```

```{r usarrests-head-solution}
head(USArrests)
```

```{r  usarrests-head-check}
lecture_grade_code()
```

We can observe the first `5` samples from the data set. You might notice that the
the first "column" contains the state names with no column name. This is the row name of the
`data.frame`. You can access the row names using `rownames` function. Try that now:

```{r usarrests-rownames, exercise=TRUE}
___(USArrests)
```

```{r usarrests-rownames-solution}
rownames(USArrests)
```

```{r  usarrests-rownames-check}
lecture_grade_code()
```

### `summary`

To get the summary of the data, you can use the `summary` function. It will give you
the summary of each column in the data set. Try that now:
```{r usarrests-summary, exercise=TRUE}
___(USArrests)
```

```{r usarrests-summary-solution}
summary(USArrests)
```

```{r  usarrests-summary-check}
lecture_grade_code()
```

### Scatter plot
If you remember, we spoke about the necessity of visualizing the data before applying
any analysis. Let us have a look on the data using scatter plot via `ggplot2` library.
Try to create a scatter plot using `ggplot2` library with `Murder` on x-axis,
`Assault` on y-axis and `UrbanPop` as the size of the points. Use `geom_point` to
create the scatter plot:
```{r usarrests-visualization, exercise=TRUE}
USArrests |> 
  ggplot(aes(x = ___, y = ___, size = ___)) +
  ___() +
  theme_bw()
```

```{r usarrests-visualization-solution}
USArrests |> 
  ggplot(aes(x = Murder, y = Assault, size = UrbanPop)) +
  geom_point() +
  theme_bw()
```

```{r usarrests-visualization-check}
lecture_grade_code()
```

## Normalization & Standardization
As you already know, the data, which are processed by the clustering algorithms
should be "somehow adjusted" in order make the values of individual features more 
comparable. The most common approaches are normalization and standardization. 

### Normalization
```{r normalization-setup, echo=FALSE}
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

standardize <- function(x) {
  x |>
    scale() |>
    as.vector()
}

usarrests_norm <- 
  USArrests |> 
  mutate(across(everything(), normalize))
```

Normalized data are scaled to the range between 0 and 1. The scaling reduces the 
effect of outliers (data deviating much from the other data). Define your own function
`normalize`, which subtracts the `min`imum value from vector `x` and then divides 
the result with the range of the original vector (`max`imum - `min`imum):

```{r normalize, exercise=TRUE}
___ <- function(___) {
  (___ - ___(___)) / (___(___) - ___(___))
}
```

```{r normalize-solution}
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```

```{r normalize-check}
lecture_grade_code()
```

Test it on the `USArrests` data set using the `tidyverse` function `mutate` 
combined with the `across` function. The `across` function allows you to apply the same
function to multiple columns. Use `everything()` to apply the function to all columns
in the data set. Store the new dataset in the variable `usarrests_norm`:
```{r usarrests-norm, exercise=TRUE, exercise.setup = "normalization-setup"}
___ <- 
  USArrests |> 
  ___(across(___(), ___))
```

```{r usarrests-norm-solution}
usarrests_norm <- 
  USArrests |> 
  mutate(across(everything(), normalize))
```

```{r usarrests-norm-check}
lecture_grade_code()
```

Check the head of the new data set using the `head` function:
```{r usarrests-norm-head, exercise=TRUE, exercise.setup = "normalization-setup"}
___(___)
```

```{r usarrests-norm-head-solution}
head(usarrests_norm)
```

```{r usarrests-norm-head-check}
lecture_grade_code()
```

You can observe that all columns are now scaled to the range between `0` and `1`.

### Standardization
Standardized data have mean value of 0 and their standard deviation is 1. For performing
the standardization we can employ z-score standardization, which is implemented 
in R with `scale` function. Try the execute function on `USArrests` `Murder` column:
```{r usarrests-scale, exercise=TRUE}
___(USArrests$___)
```

```{r usarrests-scale-solution}
scale(USArrests$Murder)
```

```{r usarrests-scale-check}
lecture_grade_code()
```

You might notice that the output looks different from what we are used to be an
output of a `data.frame`. The output is a matrix. We need to fix that in order
to work with the data more easily. Convert the matrix to a `data.frame`. Create
function `standardize`, which will scale the input vector `x` using the 
`scale` function and converts its result to a vector:

```{r standardize, exercise=TRUE}
___ <- function(x) {
  ___ |>
    ___() |>
    as.___()
}
```

```{r standardize-solution}
standardize <- function(x) {
  x |>
    scale() |>
    as.vector()
}
```

```{r standardize-check}
lecture_grade_code()
```

Test it on the `USArrests` data set again using `mutate` combined with the `across` function.
Use `everything()` to apply the function to all columns. Try that now: 
```{r usarrests-stand, exercise=TRUE, exercise.setup = "normalization-setup"}
USArrests |> 
  ___(across(___(), ___))
```

```{r usarrests-stand-solution}
USArrests |> 
  mutate(across(everything(), standardize))
```

```{r usarrests-stand-check}
lecture_grade_code()
```

## K-means clustering
The first algorithm we are going to explore is k-means, which optimizes the clusters
using two step approach over the clearly defined criteria function. 
The algorithm is implemented with the function `kmeans`, which requires two 
input parameters: the data and `k` number of clusters. Let us try to run the 
algorithm on the `USArrests` data set. With `k` set to `3`:

```{r kmeans-orig-3, exercise=TRUE}
kmeans_orig_3 <-
  ___(___, centers = ___)
```

```{r kmeans-orig-3-solution}
kmeans_orig_3 <-
  kmeans(USArrests, centers = 3)
```

```{r kmeans-orig-3-check}
lecture_grade_code()
```

Check the result stored in `kmeans_orig_3` of the clustering using the `str` function:
```{r kmeans-orig-3-str, exercise=TRUE, exercise.setup = "kmeans-orig-3-solution"}
str(___)
```

```{r kmeans-orig-3-str-solution}
str(kmeans_orig_3)
```

```{r kmeans-orig-3-str-check}
lecture_grade_code()
```

You can see that the result is stored as a list containing multiple elements. The
most important element is the `cluster`, which contains the cluster label for each
sample. The `centers` element contains the coordinates of the cluster centers.
Other elements contains the criteria function values, sizes of clusters, etc.

We can explore the clusters using the `fviz_cluster` function from the `factoextra` package.
The function requires the clustering result `kmeans_orig_3` and the original data set `USArrests`.
Try the function now:
```{r fviz-cluster, exercise=TRUE, exercise.setup = "kmeans-orig-3-solution"}
fviz_cluster(___, data = ___)
```

```{r fviz-cluster-solution}
fviz_cluster(kmeans_orig_3, data = USArrests)
```

```{r fviz-cluster-check}
lecture_grade_code()
```

You can observe 3 clusters in the plot. The points/samples are colored according to the
clustering result. The cluster centers are marked with a larger "shape" corresponding
to the cluster. Points are named because the original data contained row names. 
The plot is two dimensional projection of the multidimensional data using the first two
principal components obtained using PCA.

### Normalized data
You might notice that the clusters are not very well separated. One of the reasons
might be the fact that the data are not normalized. Let us try to run the clustering algorithm
on the normalized data `usarrests_norm` using the same `k` value. Try that now:

```{r kmeans-norm-3, exercise=TRUE, exercise.setup = "normalization-setup"}
kmeans_norm_3 <-
  ___(___, centers = ___)
```

```{r kmeans-norm-3-solution}
kmeans_norm_3 <-
  kmeans(usarrests_norm, centers = 3)
```

```{r kmeans-norm-3-check}
lecture_grade_code()
```

Check the result stored in `kmeans_norm_3` of the clustering using the `str` function:
```{r kmeans-norm-3-str, exercise=TRUE, exercise.setup = "kmeans-norm-3-solution"}
str(___)
```

```{r kmeans-norm-3-str-solution}
str(kmeans_norm_3)
```

```{r kmeans-norm-3-str-check}
lecture_grade_code()
```

Finally, you can check the clustering result using the `fviz_cluster` function. 
Call the function with the clustering result `kmeans_norm_3` and the original 
data set `usarrests_norm`:

```{r fviz-cluster-norm, exercise=TRUE, exercise.setup = "kmeans-norm-3-solution"}
fviz_cluster(___, data = ___)
```

```{r fviz-cluster-norm-solution}
fviz_cluster(kmeans_norm_3, data = usarrests_norm)
```

```{r fviz-cluster-norm-check}
lecture_grade_code()
```

We can observe that the clusters are much better separated. This is the result
of normalization, which made the features more comparable.

## Selecting `k`
When performing clustering, we need to select the number of clusters `k`. You can
do that empirically by running the clustering algorithm with different `k` values and
checking the clustering result. However, this is not very efficient, thus several 
methods can be used to determine the optimal `k` value. 

### Elbow method
The elbow method is based on the idea that the clustering algorithm will
produce a better clustering result with larger `k` values. However, the improvement
is not linear and for specific case of `k` the improvement is not significant producing
the "elbow" in the plot of the clustering objective function value.

Use the `fviz_nbclust` function to visualize the values of within-cluster sum of squares
(`wss`) for different `k` values. The function requires the data set (`usarrests_norm`),
the clustering algorithm and the specification of the measure (`wss`):

```{r elbow-method, exercise=TRUE, exercise.setup = "kmeans-norm-3-solution"}
fviz_nbclust(___, ___, method = "___") # data, algorithm, measure
```

```{r elbow-method-solution}
fviz_nbclust(usarrests_norm, kmeans, method = "wss") # data, algorithm, measure
```

```{r elbow-method-check}
lecture_grade_code()
```

You can observe two "elbows" in the dataset - one for `k = 2` and the second for 
`k = 4`. Now you can check the results for both values and decide, which holds 
better outcomes. First, train models for both `k`s using `usarrests_norm`:

```{r kmeans-norm-2-4, exercise=TRUE, exercise.setup = "normalization-setup"}
kmeans_norm_2 <-
  ___(___, centers = ___)

kmeans_norm_4 <-
  ___(___, centers = ___)
```

```{r kmeans-norm-2-4-solution}
kmeans_norm_2 <-
  kmeans(usarrests_norm, centers = 2)

kmeans_norm_4 <-
  kmeans(usarrests_norm, centers = 4)
```

```{r kmeans-norm-2-4-check}
lecture_grade_code()
```

Next we can plot the results of the clustering next to each other using `patchwork` package 
in combination with the `fviz_cluster` function. For the data use the `usarrests_norm` and
for the clustering results use `kmeans_norm_2` and `kmeans_norm_4`:

```{r fviz-cluster-norm-2-4-setup}
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

usarrests_norm <- 
  USArrests |> 
  mutate(across(everything(), normalize))

kmeans_norm_2 <-
  kmeans(usarrests_norm, centers = 2)

kmeans_norm_3 <-
  kmeans(usarrests_norm, centers = 3)

kmeans_norm_4 <-
  kmeans(usarrests_norm, centers = 4)

sil <- silhouette(kmeans_norm_3$cluster, dist(usarrests_norm))

distances <- dist(usarrests_norm)

hclust_res <- hclust(distances)

hclust_3 <- cutree(hclust_res, k = 3)

dend <- as.dendrogram(hclust_res)
```

```{r fviz-cluster-norm-2-4, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup",fig.height=10}
fviz_cluster(___, data = ___) /
  fviz_cluster(___, data = ___)
```

```{r fviz-cluster-norm-2-4-solution}
fviz_cluster(kmeans_norm_2, data = usarrests_norm) / 
  fviz_cluster(kmeans_norm_4, data = usarrests_norm)
```

```{r fviz-cluster-norm-2-4-check}
lecture_grade_code()
```

### Silhouette method
The another method for determining the number of clusters is the Silhouette score. 
Let us plot the silhouette score for our 3-means clustering result `kmeans_norm_3` we used
above. First we need to compute the silhouette score using the `silhouette` function.
It takes two arguments first are the clustering labels and the second is the distance
matrix. The distance matrix can be computed from `usarrests_norm` using the `dist` function. 
Try it now:
```{r silhouette, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
sil <- silhouette(___$cluster, dist(___)) 
```

```{r silhouette-solution}
sil <- silhouette(kmeans_norm_3$cluster, dist(usarrests_norm))
```

```{r silhouette-check}
lecture_grade_code()
```

The result is a matrix with the silhouette score for each sample. Plot the silhouette `sil`
score using the `fviz_silhouette` function. 

```{r fviz-silhouette, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
fviz_silhouette(___)
```

```{r fviz-silhouette-solution}
fviz_silhouette(sil)
```

```{r fviz-silhouette-check}
lecture_grade_code()
```

And now we can plot the silhouette overall average score based on the number of 
clusters `k`. We then need to pick the highest value and use the corresponding `k`
as the number of clusters. First plot the average silhouette score using the `fviz_nbclust` function
with `method = "silhouette"` for our normalized data set `usarrests_norm` and the 
`kmeans` algorithm:

```{r silhouette-avg, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
fviz_nbclust(___, ___, method = "___") # data, algorithm, measure
```

```{r silhouette-avg-solution}
fviz_nbclust(usarrests_norm, kmeans, method = "silhouette") # data, algorithm, measure
```

```{r silhouette-avg-check}
lecture_grade_code()
```

## Hierarchical clustering
The second method we will be working with is the Agglomerative hierarchical clustering.
In the R implementation it does not work with the data set itself, but process 
the distance (dissimilarity) object produced by `dist` function. In principle it is matrix, 
but R process the data and provides you with object containing all matrix elements. 
Compute the distance matrix using the `dist` function on the normalized data set `usarrests_norm`:
```{r dist, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
distances <- dist(___)
```

```{r dist-solution}
distances <- dist(usarrests_norm)
```

```{r dist-check}
lecture_grade_code()
```

Check the first 5 rows and columns of the `distances` matrix using the `as.matrix` function:
```{r dist-matrix, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
___(___)[1:5, 1:5]
```

```{r dist-matrix-solution}
as.matrix(distances)[1:5, 1:5]
```

```{r dist-matrix-check}
lecture_grade_code()
```

You can observe that the distance matrix is symmetric and the diagonal contains
`0`s. The distance matrix is used as an input for the clustering algorithm. You 
can perform the clustering using the `hclust` function with the `distances` matrix. 
Try that now:
```{r hclust, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
hclust_res <- hclust(___)
```

```{r hclust-solution}
hclust_res <- hclust(distances)
```

```{r hclust-check}
lecture_grade_code()
```

### Dendrogram
Now we can plot the resulting object `hclust_res` using the `plot` function. The plot is a
dendogram, which shows the hierarchical structure of the clustering. Plot the
dendogram:
```{r hclust-plot, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
plot(___)
```

```{r hclust-plot-solution}
plot(hclust_res)
```

```{r hclust-plot-check}
lecture_grade_code()
```

Finally, we can cut the dendogram into `k` clusters using the `cutree` function.
Cut the dendogram into `3` clusters using the `hclust_res` object:

```{r hac-cut, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
hclust_3 <- cutree(___, k = ___)
```

```{r hac-cut-solution}
hclust_3 <- cutree(hclust_res, k = 3)
```

```{r hac-cut-check}
lecture_grade_code()
```

Let us check the resulting cut via calling the cutted tree stored in `hclust_3`:
```{r hac-cut-view, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
___
```

```{r hac-cut-view-solution}
hclust_3
```

```{r hac-cut-view-check}
lecture_grade_code()
```

### Pretty dendogram

To "prettify" the dendrogram with our clusters we can employ the functions from `dendextend` 
package. First the resulting object `hclust_res` needs to be converted to the dendrogram object
of the `dendextend` package. Convert it using `as.dendrogram` function:
```{r hac-dendextend, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
dend <- as.dendrogram(___)
```

```{r hac-dendextend-solution}
dend <- as.dendrogram(hclust_res)
```

```{r hac-dendextend-check}
lecture_grade_code()
```

Now we can color the branches of the dendrogram using the `color_branches` function and
plot it using the `plot` function. Set the `k` parameter to `3`. Do that now:
```{r hac-dendextend-plot, exercise=TRUE, exercise.setup = "fviz-cluster-norm-2-4-setup"}
color_branches(dend, k = ___) |>
  ___()
```

```{r hac-dendextend-plot-solution}
color_branches(dend, k = 3) |>
  plot()
```

```{r hac-dendextend-plot-check}
lecture_grade_code()
```

## Conclusion
This concludes our third seminar lecture, where you learned about clustering methods
`kmeans` and `hclust`. Next comes the third assignment, which will available after of our seminar.

Here we repeat the important information regarding your consent to the data collection. The following hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:

```{r, echo=FALSE}
textOutput("id_display2")
```
\
Please, keep it for your records in case you change your mind and want to delete your data.
