---
title: "Unsupervised Learning - assignment"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE, context="server"}
library(learnr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

library(readr)
library(tidyverse)
library(ggplot2)
library(factoextra)
library(dendextend)
library(cluster)
library(patchwork)
```

## Assigment

In our third assignment we will be working with the subset of (OULAD data set)[https://analyse.kmi.open.ac.uk/open_dataset]. 
This data set has been released by the Open University in United Kingdom and 
contains the information about approximately 30,000 students of the university in
years 2013 and 2014. In our case, we will be using the assessment data from the one 
course. The assignment data are in the csv format (as usual) and stored data set
contains 4 columns corresponding to the 4 assignments (A1 - A4). The values in cells
are achieved scores in assignments by individual students. Each row represents one student.

```{r exercise-setup, echo=FALSE}
input_data <- read_csv("data/assigment_data.csv")
```

### Check data   
Explore the `input_data` using `head`, `str` and `summary` functions.
```{r exercise-1, exercise=TRUE, exercise.setup = "exercise-setup"}
___(___)

___(___)

___(___)
```

### Visualize

First visualize data using `ggplot2` library in combination with `patchwork` package.

#### Histograms
Create histograms of individual features. For plotting use `geom_histogram` function
in combination with `facet_wrap` function, which will create a grid of plots. Data 
are stored in `input_data` variable. To use the `facet_wrap` function you need to
reshape the data from wide to long format. Use `pivot_longer` function from `tidyverse` package.
For selection of all columns use `everything()` function. Create histogram now:

```{r exercise-2, exercise=TRUE, exercise.setup = "exercise-setup"}
# prepare data
___ |> 
  ___(___ = ___(), 
      ___ = "assignment",
      ___ = "score") |>
  # plot
  ___(___(___ = score)) +
  ___() +
  ___(~___) # grid based on the assignment
```

#### Scatter plots
Now create scatter plots of the data. Use `geom_point` function to create scatter plots
of A1 vs A2 and A3 vs A4. Use `patchwork` package to combine the plots using the `/` operator.
Store individual scatter plots in `g1` and `g2` variables before combining them with `/` operator.

```{r exercise-3, exercise=TRUE, exercise.setup = "exercise-setup"}
# A1 vs A2
g1 <-
  ___ |>
  ___(___(___ = ___, ___ = ___)) +
  ___()

# A3 vs A4
g2 <-
  ___ |>
  ___(___(___ = ___, ___ = ___)) +
  ___()

# combine plots and plot
___ / ___
```

### Imputation

When checking the data you might have noticed that some values are missing. Now it is 
the time to replace `NA` values with `0`. Replace them now and store the result in
`data_imputed` variable:

```{r exercise-4, exercise=TRUE, exercise.setup = "exercise-setup"}
# copy  input_data to new variable
___ <- ___

# replace NA values with 0 in data_imputed
___[___(___)] <- ___
```

Hint: There exist function `is.na`.

```{r exercise-2-setup, echo=FALSE}
input_data <- read_csv("data/assigment_data.csv")

data_imputed <- input_data
data_imputed[is.na(data_imputed)] <- 0

normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```

### Normalize
You have now the idea how the data looks like and you replaced all missing values
with `0`. Normalize them in order to get ready for the clustering. 
Use function `normalize` from the lecture:
```{r normalize, eval = FALSE}
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
```
Apply the function to all columns of the `data_imputed` data, store the result 
in `data_normalized` variable:

```{r exercise-5, exercise=TRUE, exercise.setup = "exercise-2-setup"}
(
___ <- 
  ___ |>
  ___(___(___(), ___))
)
```

> Note: you might notice extern `()` in the code. This trick forces the R to print
the result of the operation to the console. Otherwise, the result is not printed
to the console.

### Pick optimal `k` using Elbow method
```{r exercise-3-setup, echo=FALSE}
input_data <- read_csv("data/assigment_data.csv")

data_imputed <- input_data
data_imputed[is.na(data_imputed)] <- 0

normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

data_normalized <- data_imputed |>
  mutate(across(everything(), normalize))
```

Finally, your data are ready for clustering using `kmeans` algorithm. At first,
pick the optimal number of clusters `k` using Elbow method. Use `fviz_nbclust` function
to produce the plot, which will help you with selection. Use `data_normalized` for
the clustering. The `method` parameter should reflect the Elbow method using total
within-cluster sum of squares (`wss`).

```{r exercise-6, exercise=TRUE, exercise.setup = "exercise-3-setup"}
___(___, 
    ___, 
    ___ = "___") 
```

### k-means

With picked number of `centers` from the previous step. Do the clustering of the `data_normalized`
using `kmeans` function. Store results in `kmeans_result` variable.

```{r exercise-7, exercise=TRUE, exercise.setup = "exercise-3-setup"}
___ <- 
  ___ |>
  ___(___ = ___)
```

### Visualization of result
Finally, visualize the clustering result using `fviz_cluster` function. Because
I cannot know, which `k` you will pick, we need to copy the code from previous step into this
code "chunk". Then write the code for visualization.

```{r exercise-8, exercise=TRUE, exercise.setup = "exercise-3-setup"}
# from previous step
___ <- 
  ___ |>
  ___(___ = ___)

# visualization
___(___, ___ = ___)
```

### Submission 
Now, you have the final result. Copy your code from the last step (*Visualization of result*) and paste it 
to the Moodlelab in the [third assignment](https://moodlelab.hu-berlin.de/mod/assign/view.php?id=11433).
