---
title: "Supervised methods"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/custom.css
runtime: shiny_prerendered
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(glue)
library(dplyr)
library(curl)
library(jsonlite)
library(openssl)
library(tidyverse)
library(caret)
library(patchwork)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

options(user_id = NULL)

lecture_grade_code <- function() {
  grade_this_code(
    correct = "Correct!",
    incorrect = "Try again. {code_feedback()}"
  )
}
```

```{js, context = "head", echo = FALSE}
// JavaScript code to generate a unique ID and send it to R
$(document).ready(function() {
  // Function to send the unique ID to R
  function sendIDToR(uniqueID) {
    // Store in localStorage for persistence
    localStorage.setItem('learnr_user_id', uniqueID);
    
    // Send to R via Shiny if available
    if (typeof Shiny !== 'undefined' && typeof Shiny.setInputValue === 'function') {
      Shiny.setInputValue('user_unique_id', uniqueID);
      console.log('IP address sent to R:', uniqueID);
    } else {
      // If Shiny isn't ready yet, try again in 500ms
      setTimeout(function() { sendIDToR(uniqueID); }, 500);
    }
  }
  
  // Check if we already have the ID in the localStorage
  const uniqueID = localStorage.getItem('learnr_user_id');
  if (uniqueID) {
    // If we have it, send it to R
    sendIDToR(uniqueID);
  } else {
    // Generate a new unique ID
    sendIDToR(Math.random().toString(36).substring(2, 15));
  }
});
```

```{r, context="server", echo=FALSE}
# IMPORTANT: This chunk runs in the Shiny server context
# Here 'input' object is available

# Create a reactive value to store the user's IP address
user_id <- reactiveVal(NULL)

# Observer that captures the IP address from JavaScript
observeEvent(input$user_unique_id, {
  # Store the IP address in our reactive value
  user_id(md5(input$user_unique_id))
  
  # Also store in options for global access
  id_data <- getOption("user_id")
  if (is.null(id_data)) {
    id_data <- md5(input$user_unique_id)
    options(user_id = id_data)
  }
})

# For debugging: Display the captured IP address
output$id_display <- renderText({
  if(is.null(user_id())) {
    "ID not yet generated"
  } else {
    user_id()
  }
})

output$id_display2 <- renderText({
  if(is.null(user_id())) {
    "ID not yet generated"
  } else {
    user_id()
  }
})

Sys.sleep(0.7)

event_recorder <- function(tutorial_id, 
                           tutorial_version,
                           user_id,
                           event, 
                           data) {
  REPO <- "/Users/jakubkuzilek/Documents/projects/temp"
  
  HASH <- getOption("user_id")
  
  if (file.exists(glue("{REPO}/data_{HASH}.rds"))) {
    t <- readRDS(glue("{REPO}/data_{HASH}.rds"))
  } else {
    t <- 
      tibble(
        time = .POSIXct(numeric(0)),
        tutorial_id = character(),
        tutorial_version = character(),
        user_id = HASH,
        hash = character(),
        event = character(),
        data = list()
      )
  }
  
  t <- bind_rows(t, 
                 tibble(
                   time = Sys.time(),
                   tutorial_id = tutorial_id,
                   tutorial_version = tutorial_version,
                   user_id = HASH,
                   event = event,
                   data = list(data)
                 )
  )
  
  saveRDS(t, file = glue("{REPO}/data_{HASH}.rds"))
}

options(tutorial.event_recorder = event_recorder)
```

## Introduction
At the end of this lecture you will:

- know how to implement k-Nearest Neighbours algorithm (k-NN)
- be able to conduct ML experiment using `caret` package

> For simplicity, we will not be using the R and RStudio installed on your computer. We will be
using interactive excercises created using [learnr](https://cran.r-project.org/web/packages/learnr/index.html) package.

> When you find `___` in the code chunk, it means that you need to fill in the missing part of the code.

## Informed Consent
This informed consent form is for a project investigating the interactions with the online
tutorial platform created using `learnr` package.

Name of Investigator: Dr. Jakub Kuzilek (jakub.kuzilek@hu-berlin.de)

Name of Supervisor: Prof. Dr. Niels Pinkwart (pinkwart@hu-berlin.de)

Name of Institution: Humboldt-Universität zu Berlin

**You can download this informed consent in the Moodle course.**

**The data from multiple tutorials will be collected and analysed.The informed consent is the same. For each session, you will get hash, which is the only unique identifier of your interactions with the material.** 

### Information sheet

#### Introduction of investigators
Jakub Kuzilek is a researcher in fields of educational data mining and learning analytics since completion of his PhD in Artificial Intelligence and Biocybernetics at the Czech Technical University. In 2020 he joined Prof. Dr. Pinkwarts's research group and currently working on the project IMPACT.

#### Purpose of the study
The purpose of this study is to investigate the interactions with the online tutorial platform created using the `learnr` package. The data will be used to improve the design of the tutorial and to understand how users interact with the platform.

#### Type of Research Intervention
To support and investigate these learning and interaction processes interaction data will be used. In addition performance data of the students will be collected.

#### Research Data Management
The collected data will be anonymized and merged. The collected data are stored on the servers of the Computer and Media Service (CMS) of the Humboldt-Universität zu Berlin in Germany and the computers of the researchers. The data access is protected and regulated via a login and the access is restricted to the members of the project team. The personal data will be stored on secured location accessible via authenticated encrypted connection (VPN and SSH). The data are archived for another 10 years after the end of the project according to good scientific practice. The anonymized data may be published to enable Open Access for the general public. The participant may gain access to his/her recorded data, request a correction to his/her personal data record, or withdraw from the research study by contacting the Investigator:
Jakub Kuzilek at jakub.kuzilek@hu-berlin.de

In case of doubt as to the lawfulness of the processing of personal data, data subjects have the right to lodge a complaint with a Member State supervisory authority. The competent supervisory authority for Humboldt-Universität zu Berlin is Die Berliner Beauftragte für Datenschutz und Informationsfreiheit, Friedrichstraße 219, 10969 Berlin. The official data protection officer for Humboldt-Universität zu Berlin is: Gesine Hoffmann-Holland, Tel: +49 (30) 2093-20022, datenschutz@uv.hu-berlin.de.

#### Voluntary participation
Participation in this research project is entirely voluntary. It is up to the participant to decide whether or not to participate and there is no explicit or implicit obligation to participate. The participant has the possibility to cancel the participation at any time - even after the signing of the Informed Consent - without any disadvantages for the participant. 

#### Documentation of the consent
One copy of this document will be kept separately from the research records of this study using a pseudonym id in order to remove data after the request from the participant. The contact details of the participants will be deleted after the project is finished. To assure data deletion of the requesting subject a unique identifier in form of the email address and is required.

### Certificate of consent
- I agree to the collection of my pseudonymized interaction data in the course "Machine Learning for Education".
- I agree to the collection of my pseudonymized perfomance data in the course "Machine Learning for Education".
- I agree to have the anonymized interaction data and the anonymized performance data published under the Creative Commons Attribution 4.0 International (CC-BY) Licence (https://creativecommons.org/licenses/ by/4.0/) and so I allow it to be used by other people for other projects. 
- I agree to have the interaction data, and the performance data archived for 10 years after finishing of the project.

This hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:

```{r, echo=FALSE}
textOutput("id_display")
```
\
Please, keep it for your records in case you change your mind and want to delete your data.

```{r informed-consent, echo=FALSE}
question_radio("I agree to the collection of my pseudonymized interaction data in the course 'Machine Learning for Education', lecture: 'Supervised Methods'.", 
               correct = NULL, 
               answer("yes", correct = TRUE),
               answer("no", correct = TRUE)
)
```

## Reproducibility
The k-NN is not influenced by the initialization problem like k-means algorithm, 
but when performing the ML experiment you will be splitting data into the folds.
Thus you will never get the same result. You need to "seed" properly random number
generator to obtain the stable and reproducible results. This will guarantee that your 
results are stable between multiple runs. Set the seed to the `1234` like in 
previous lecture:
```{r set-seed, exercise=TRUE}
___(___)
```

```{r set-seed-solution}
set.seed(1234)
```

```{r set-seed-check}
lecture_grade_code()
```

## Data
In this lecture we will be using data set called `iris`. You can read details about dataset using `?iris`. 
It contains 150 samples of three different species of iris and measurements of their sepal (length \& width) 
and petal (length \& width) leafs. At first, check the first 10 entries of dataset using `head()` function:
```{r iris-head, exercise=TRUE}
___(___,___)
```

```{r iris-head-solution}
head(iris, 10)
```

```{r iris-head-check}
lecture_grade_code()
```

### Missing values
When you are going to perform machine learning training on the dataset you will
need to check if some data samples have missing values. You can check it using 
`summary()` function. This function will show you the minimum, maximum, mean, median and
quantiles of the numeric variables and number of unique values in the categorical variables. It
will also shows you number of `NA`s, which is the value that corresponds to missing value in R.
Try it:

```{r iris-summary, exercise=TRUE}
___(___)
```

```{r iris-summary-solution}
summary(iris)
```

```{r iris-summary-check}
lecture_grade_code()
```

### Visual exploration
Good. You have no missing values in the dataset. So it is time to explore it visually.
You are already familiar with `ggplot2` package, so you can use it to create the scatter plot (`geom_point`) of the dataset.
Plot the scatterplot with `Sepal.Length` on the x-axis and `Sepal.Width` on the y-axis. Assign
the `Species` variable to the `color` aesthetic to see the distribution of samples.

```{r iris-ggplot, exercise=TRUE}
___(___,
    ___(___ = ___, 
        ___ = ___, 
        ___ = ___)
    ) +
  ___()
```

```{r iris-ggplot-solution}
ggplot(iris, 
       aes(x = Sepal.Length, 
           y = Sepal.Width, 
          color = Species)
       ) +
  geom_point()
```

```{r iris-ggplot-check}
lecture_grade_code()
```

### 

You can observe that there is clear difference of the setosa class compared
to versicolor and virginica species. We can also observe that versicolor is somewhere
in the middle "bellow" the virginica, but the clear distinction is not possible.

## k-Nearest Neighbours

The k-Nearest Neighbors (k-NN) measures the distance between new sample to the 
samples in the training set, picks k nearest samples and classifies the new sample
according to majority voting of the classes of the k nearest samples. We will 
implement simple version of the k-NN model with `iris` data set. 

### New data 
At first, create new sample of iris having 
`Sepal.Length = 5.0, Sepal.Width = 3.5, Petal.Length = 1.5, Petal.Width = 0.2` 
and `Species = "setosa"`. Create new `date.frame` containing just one row with the 
values of the new sample:
```{r new-sample, exercise=TRUE}
new_sample <- ___(___ = ___, 
                   ___ = ___, 
                   ___ = ___, 
                   ___ = ___, 
                   ___ = "setosa")
```

```{r new-sample-solution}
new_sample <- data.frame(Sepal.Length = 5.0, 
                          Sepal.Width = 3.5, 
                          Petal.Length = 1.5, 
                          Petal.Width = 0.2, 
                          Species = "setosa")
```

```{r new-sample-check}
lecture_grade_code()
```

### 

Let us visually check the sample we created. Use `ggplot2` to plot the scatterplot
with `Sepal.Length` on the x-axis and `Sepal.Width` on the y-axis. First, plot
the `iris` data set using `geom_point()` function and store it as `g` variable. 
Color the points by `Species`. Try it now:
```{r new-sample-ggplot, exercise=TRUE}
(
g <-
  ___(___, 
      aes(___ = ___, 
          ___ = ___, 
          ___ = ___)
       ) +
  ___()
)
```

```{r new-sample-ggplot-solution}
(
g <-
  ggplot(iris, 
      aes(x = Sepal.Length, 
          y = Sepal.Width, 
          color = Species)
       ) +
  geom_point()
)
```

```{r new-sample-ggplot-check}
lecture_grade_code()
```

Notice the outer brackets, which forces the `g` variable to be called after the 
creation. This is useful for saving some code writing.

###

Now, add new sample to the plot using `geom_point()` function. Set the `data` argument
to `new_sample` and set the `x` and `y` aesthetics to `Sepal.Length` and `Sepal.Width`.
To better find the point in the plot set the `color` argument to `"black"` and `size` to `5`:
```{r new-sample-ggplot2-setup}
g <-
  ggplot(iris, 
      aes(x = Sepal.Length, 
          y = Sepal.Width, 
          color = Species)
       ) +
  geom_point()

new_sample <- data.frame(Sepal.Length = 5.0, 
                          Sepal.Width = 3.5, 
                          Petal.Length = 1.5, 
                          Petal.Width = 0.2, 
                          Species = "setosa")
```

```{r new-sample-ggplot2, exercise=TRUE}
g +
  ___(___ = ___, 
      ___(___ = ___, 
          ___ = ___), 
          ___ = "___", 
          ___ = ___)
```

```{r new-sample-ggplot2-solution}
g +
  geom_point(data = new_sample, 
             aes(x = Sepal.Length, 
                 y = Sepal.Width), 
             color = "black", 
             size = 5)
```

```{r new-sample-ggplot2-check}
lecture_grade_code()
```

### Distance
Great! So we have our new sample and our `iris` dataset. Now, we need to calculate
the distance between the new sample and all samples in the `iris` dataset. There 
are few steps to do that. At first, append the new sample to the `iris` dataset 
and remove the `Species` column from the new dataset (`new_iris`), because we will not use that 
for distance calculation. For appending the row to the `data.frame` use 
function `bind_rows()`. For selecting the columns use `select()` function. Do it now:
```{r new-sample-iris-setup}
new_sample <- data.frame(Sepal.Length = 5.0, 
                          Sepal.Width = 3.5, 
                          Petal.Length = 1.5, 
                          Petal.Width = 0.2, 
                          Species = "setosa")
```

```{r new-sample-iris, exercise=TRUE}
new_iris <- 
  ___ |> # add sample to the first position
  ___(___) |>
  ___(-___) 
```

```{r new-sample-iris-solution}
new_iris <- 
  new_sample |> # add sample to the first position
  bind_rows(iris) |>
  select(-Species) 
```

```{r new-sample-iris-check}
lecture_grade_code()
```

### 

Awesome! Now we have our new sample in the `new_iris` dataset. Let us proceed to 
calculating the distance between the new sample and all samples using the 
`dist()` function. The function returns numeric vector, which is hard to process.
Thus convert the vector to the matrix and store it in `distances` variable:
```{r new-sample-distances-setup}
new_sample <- data.frame(Sepal.Length = 5.0, 
                          Sepal.Width = 3.5, 
                          Petal.Length = 1.5, 
                          Petal.Width = 0.2, 
                          Species = "setosa")

new_iris <- 
  new_sample |> # add sample to the first position
  bind_rows(iris) |>
  select(-Species) # drop labels
```

```{r new-sample-distances, exercise=TRUE}
___ <- 
  ___ |>
  ___() |>
  ___()
```

```{r new-sample-distances-solution}
distances <- 
  new_iris |>
  dist() |>
  as.matrix()
```

```{r new-sample-distances-check}
lecture_grade_code()
```

### Selecting the nearest neighbours
Now, we have all distances between all samples computed. Thus, we need to pick
the samples closest to the new sample. Select the 1st row of the distances matrix
to get the distances of our new sample to all other samples. Then, sort the distances
and select the first `k = 3` samples. Use `order()` function to sort the distances and
`head()` to extract the first `k` samples. Store the result in `nearest` variable:
```{r new-sample-nearest-setup}
new_sample <- data.frame(Sepal.Length = 5.0, 
                          Sepal.Width = 3.5, 
                          Petal.Length = 1.5, 
                          Petal.Width = 0.2, 
                          Species = "setosa")

new_iris <-
  new_sample |> # add sample to the first position
  bind_rows(iris) |>
  select(-Species) # drop labels

distances <-
  new_iris |>
  dist() |>
  as.matrix()
```

```{r new-sample-nearest, exercise=TRUE}
___ <- 
  ___[___,] |>
  ___() |>
  ___()
```

```{r new-sample-nearest-solution}
nearest <- 
  distances[1,] |>
  order() |>
  head(3)
```

```{r new-sample-nearest-check}
lecture_grade_code()
```

### 

Finally, we do know our nearest neighbours, but we do not know their labels. So 
we need to extract the labels from the `Species` column of the `iris` dataset,
select the labels of the nearest neighbours, count the labels and select the most frequent one.
To do that you need to `filter` the `iris` dataset by the `row_number`s of the `nearest` neighbours,
then `count` the labels (`Species` column), `arrange` them in `desc`ending order based on the cou`n`ted values
and finally `slice` the first one. Store the value in the `classes` variable. Try it now:
```{r new-sample-nearest-labels-setup}
new_sample <- data.frame(Sepal.Length = 5.0, 
                          Sepal.Width = 3.5, 
                          Petal.Length = 1.5, 
                          Petal.Width = 0.2, 
                          Species = "setosa")
new_iris <-
  new_sample |> # add sample to the first position
  bind_rows(iris) |>
  select(-Species) # drop labels

distances <-
  new_iris |>
  dist() |>
  as.matrix()

nearest <-
  distances[1,] |>
  order() |>
  head(3)
```

```{r new-sample-nearest-labels, exercise=TRUE}
___ <- 
  ___ |>
  ___(___() %in% ___) |>
  ___(___) |>
  ___(___(___)) |>
  ___(1)
```

```{r new-sample-nearest-labels-solution}
classes <- 
  iris |>
  filter(row_number() %in% nearest) |>
  count(Species) |>
  arrange(desc(n)) |>
  slice(1)
```

```{r new-sample-nearest-labels-check}
lecture_grade_code()
```

### 

And finally, we can `print` the class of the new sample stored in `classes$Species`:
```{r new-sample-class-setup}
new_sample <- data.frame(Sepal.Length = 5.0, 
                          Sepal.Width = 3.5, 
                          Petal.Length = 1.5, 
                          Petal.Width = 0.2, 
                          Species = "setosa")

new_iris <-
  new_sample |> # add sample to the first position
  bind_rows(iris) |>
  select(-Species) # drop labels

distances <-
  new_iris |>
  dist() |>
  as.matrix()

nearest <-
  distances[1,] |>
  order() |>
  head(3)

classes <-
  iris |>
  filter(row_number() %in% nearest) |>
  count(Species) |>
  arrange(desc(n)) |>
  slice(1)
```

```{r new-sample-class, exercise=TRUE}
___(___)
```

```{r new-sample-class-solution}
print(classes$Species)
```

```{r new-sample-class-check}
lecture_grade_code()
```

## `caret` package
We tried to classify the new sample using the k-NN algorithm. But when working on 
real data you need something more robust and of more quality. For that purpose
the `R` does have wide range of algorithms implemented. Unfortunately, they all
have different interfaces and it is hard to use them together. To solve this problem
the `caret` package was created. It provides unified interface to many algorithms
and ease the machine learning workflow. 

### Data
In our example we will be again using the `iris` dataset. But this time we will
merge two species together to create a binary classification problem. Create
new dataset called `iris_caret` containing column `Species` with only two classes `virginica` and `not virginica`.
For conditional `mutate` of the data in the column use `if_else()` function. Try that now:
```{r iris-caret, exercise=TRUE}
___ <- 
  ___ |>
  ___(___ = ___(___ == "___", 
                "___", 
                "___")
      )
```

```{r iris-caret-solution}
iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )
```

```{r iris-caret-check}
lecture_grade_code()
```

### Data & setup
When you are performing the experiment with small data sets it is usually good 
approach to use [cross-validation](https://en.wikipedia.org/wiki/Cross-validation_(statistics)) for estimating the method error.
So for start we create 10 `folds`, to which we will split the data using `createFolds()` on the `Species` column:
```{r iris-caret-folds-setup}
set.seed(1234)

iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )
```

```{r iris-caret-folds, exercise=TRUE}
___ <- 
  ___(___$___, 
      list = FALSE) # return matrix instead of list
```

```{r iris-caret-folds-solution}
folds <- 
  createFolds(iris_caret$Species, 
              list = FALSE) # return matrix instead of list
```

```{r iris-caret-folds-check}
lecture_grade_code()
```

### 

Next, we need to tell `caret` "engine" how it will be training
the algorithm. This is done by creating `trainControl` object containing our ML 
experiment configuration. In this case we will use `10`-fold cross-validation (`cv`).
Store the object in `fit_control` variable:
```{r iris-caret-train-control, exercise=TRUE}
___ <- ___(method = "__", 
           number = ___) 
```

```{r iris-caret-train-control-solution}
fit_control <- trainControl(method = "cv", 
                            number = 10)
```

```{r iris-caret-train-control-check}
lecture_grade_code()
```

###

Finally, we just need to create structure for storing the results of the experiment.
Create empty `data.frame` `resutls` containing columns `fold`, `accuracy`, `precision` and `recall`:
```{r iris-caret-results, exercise=TRUE}
___ <- ___(___ = numeric(0)
           ___ = numeric(0),
           ___ = numeric(0),
           ___ = numeric(0))
```

```{r iris-caret-results-solution}
results <- data.frame(fold = numeric(0),
                       accuracy = numeric(0),
                       precision = numeric(0),
                       recall = numeric(0))
```

```{r iris-caret-results-check}
lecture_grade_code()
```

### Model training and error estimation
Finally, we can conduct our experiment. There are several elements of what we need
to do:

1. Iterate over the `folds` and train the model on data from all other folds.
2. Use the model to predict the `Species` variable on the `validation` dataset.
3. Estimate the error using our selected fold.
4. Store the results and calculate average accuracy, precision and recall.

### Iterate over the `folds`
At first we need to create the `for`-loop to iterate over the folds created previously.
In the for-loop iterate from `1` to `max(folds)`, which represents the number of folds.
Then store the `indx` of samples, which are in the corresponding fold. For that use 
`which()` function to get the indices of the samples in the fold. Finally create
`validation` and `training` datasets from the `iris_caret` dataset using the `indx` variable:
```{r caret-step-1-setup}
set.seed(1234)

iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )

folds <-
  createFolds(iris_caret$Species, 
              list = FALSE) # return matrix instead of list

fit_control <- trainControl(method = "cv", 
                            number = 10)

results <- data.frame(fold = numeric(0),
                       accuracy = numeric(0),
                       precision = numeric(0),
                       recall = numeric(0))
```

```{r caret-step-1, exercise=TRUE, exercise.lines = 30}
___(n in ___:___(___)) {
 
  ___ <- ___(___ == n) 
  ___ <- ___[___,] # use fold as a validation data set
  ___ <- ___[-___,] # use rest of data for training 
}
```

```{r caret-step-1-solution}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
}
```

```{r caret-step-1-check}
lecture_grade_code()
```

### Train the model

Next, we need to include the model training in the loop. We will use `train()` 
function from `caret` package to train the k-NN model. The function takes several arguments:
- prediction formula - the special R construct, which tells caret to predict `Species` variable using all other variables (denoted by `.`): `Species ~ .`
- `method` - the method we want to use for training, in this case `knn`.
- `data` - the `training` dataset we created in the previous step.
- `trControll` - the control object `fit_control` we created earlier to specify the training method.

Include the training now:
```{r caret-step-2-setup}
set.seed(1234)

iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )

folds <-
  createFolds(iris_caret$Species, 
              list = FALSE) # return matrix instead of list

fit_control <- trainControl(method = "cv", 
                            number = 10)

results <- data.frame(fold = numeric(0),
                       accuracy = numeric(0),
                       precision = numeric(0),
                       recall = numeric(0))
```

```{r caret-step-2, exercise=TRUE, exercise.lines = 30}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  ___ <- ___(___, # predict Species using all (.) features
             method = "___",
             data = ___,
             trControl = ___)
}
```

```{r caret-step-2-solution}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
}
```

```{r caret-step-2-check}
lecture_grade_code()
```

### Predict the `Species` variable
So the model is ready (trained and tuned using cross-validation) and now deploy it to predict 
the `Species` variable on the `validation` dataset. Use `predict()` function to do that using
the `fit` model and `validation` dataset. Store the result in `predictions` variable:
```{r caret-step-3-setup}
set.seed(1234)

iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )

folds <-
  createFolds(iris_caret$Species, 
              list = FALSE) # return matrix instead of list

fit_control <- trainControl(method = "cv", 
                            number = 10)

results <- data.frame(fold = numeric(0),
                       accuracy = numeric(0),
                       precision = numeric(0),
                       recall = numeric(0))
```

```{r caret-step-3, exercise=TRUE, exercise.lines = 30}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  ___ <- ___(___, newdata = ___) # predict the class for validation dataset
}
```

```{r caret-step-3-solution}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  predictions <- predict(fit, newdata = validation) # predict the class for validation dataset
}
```

```{r caret-step-3-check}
lecture_grade_code()
```

### Estimate the error

Hurray, we got our predictions for the validation dataset. Now we need to estimate 
the error of our model. For error estimation we will use `confusionMatrix()` function
from `caret` package. It compares the predicted values with actual values and
computes the all possible evaluation measures. The function takes three arguments:
- `data` - the predicted values, in our case `predictions`
- `reference` - the actual values, in our case `validation$Species`
- `positive` - the positive class, in our case `"virginica"`.

Note that the values needs to be factors, so you need to convert them to factors using `as.factor()`.
Store the result in `cm` variable. Continue extending your loop:
```{r caret-step-4-setup}
set.seed(1234)

iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )

folds <-
  createFolds(iris_caret$Species, 
              list = FALSE) # return matrix instead of list

fit_control <- trainControl(method = "cv", 
                            number = 10)
results <- data.frame(fold = numeric(0),
                       accuracy = numeric(0),
                       precision = numeric(0),
                       recall = numeric(0))
```

```{r caret-step-4, exercise=TRUE, exercise.lines = 30}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  predictions <- predict(fit, newdata = validation) # predict the class for validation dataset
  
  ___ <- ___(data = ___, 
             reference = as.factor(___$___), 
             positive = "___") # obtain the measurements
}
```

```{r caret-step-4-solution}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  predictions <- predict(fit, newdata = validation) # predict the class for validation dataset
  
  cm <- confusionMatrix(data = predictions, 
                        reference = as.factor(validation$Species), 
                        positive = "virginica") # obtain the measurements
}
```

```{r caret-step-4-check}
lecture_grade_code()
```

### Store the results
Finally, we need to store the results of our experiment. We will store the fold number,
accuracy, precision and recall in the `results` data frame we created earlier. 
Their values are hidden in `cm` object. The accuracy is stored in `cm$overall["Accuracy"]`,
precision is stored in `cm$byClass["Pos Pred Value"]` and recall is stored in `cm$byClass["Sensitivity"]`.
Extend the loop for the last time to store the results in `results` data frame:
```{r caret-step-5-setup}
set.seed(1234)

iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )

folds <-
  createFolds(iris_caret$Species, 
              list = FALSE) # return matrix instead of list

fit_control <- trainControl(method = "cv", 
                            number = 10)

results <- data.frame(fold = numeric(0),
                       accuracy = numeric(0),
                       precision = numeric(0),
                       recall = numeric(0))

```

```{r caret-step-5, exercise=TRUE, exercise.lines = 30}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  predictions <- predict(fit, newdata = validation) # predict the class for validation dataset
  
  cm <- confusionMatrix(data = predictions, 
                        reference = as.factor(validation$Species), 
                        positive = "virginica") # obtain the measurements
  
  # store results
  ___[n, ] <- c(n,
                ___$___["Accuracy"],
                ___$byClass["Pos Pred Value"], # = Precision
                ___$byClass["Sensitivity"]) # = Recall
}

```

```{r caret-step-5-solution}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  predictions <- predict(fit, newdata = validation) # predict the class for validation dataset
  
  cm <- confusionMatrix(data = predictions, 
                        reference = as.factor(validation$Species), 
                        positive = "virginica") # obtain the measurements
  
  # store results
  results[n, ] <- c(n,
                    cm$overall["Accuracy"],
                    cm$byClass["Pos Pred Value"], # = Precision
                    cm$byClass["Sensitivity"]) # = Recall
}

```

```{r caret-step-5-check}
lecture_grade_code()
```

### Average results
Finally, we can calculate the average accuracy, precision and recall of our model.
To do that we will use our well known `summarise()` function from `dplyr` package 
over the `results` data frame. Try it now:
```{r caret-step-6-setup}
set.seed(1234)

iris_caret <- 
  iris |>
  mutate(Species = if_else(Species == "virginica", 
                           "virginica", 
                           "not virginica")
         )

folds <-
  createFolds(iris_caret$Species, 
              list = FALSE) # return matrix instead of list

fit_control <- trainControl(method = "cv", 
                            number = 10)

results <- data.frame(fold = numeric(0),
                       accuracy = numeric(0),
                       precision = numeric(0),
                       recall = numeric(0))

```

```{r caret-step-6, exercise=TRUE, exercise.lines = 30}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  predictions <- predict(fit, newdata = validation) # predict the class for validation dataset
  
  cm <- confusionMatrix(data = predictions, 
                        reference = as.factor(validation$Species), 
                        positive = "virginica") # obtain the measurements
  
  # store results
  results[n, ] <- c(n,
                    cm$overall["Accuracy"],
                    cm$byClass["Pos Pred Value"], # = Precision
                    cm$byClass["Sensitivity"]) # = Recall
}

___ |>
  ___(___ = mean(___),
      ___ = mean(___),
      ___ = mean(___))

```

```{r caret-step-6-solution}
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  predictions <- predict(fit, newdata = validation) # predict the class for validation dataset
  
  cm <- confusionMatrix(data = predictions, 
                        reference = as.factor(validation$Species), 
                        positive = "virginica") # obtain the measurements
  
  # store results
  results[n, ] <- c(n,
                    cm$overall["Accuracy"],
                    cm$byClass["Pos Pred Value"], # = Precision
                    cm$byClass["Sensitivity"]) # = Recall
}

results |>
  summarise(accuracy = mean(accuracy),
            precision = mean(precision),
            recall = mean(recall))

```

```{r caret-step-6-check}
lecture_grade_code()
```

## Conclusion
Great! You have successfully implemented the k-NN algorithm using `caret` package and
estimated the error of the model using cross-validation. You can observe that the results
are very well, yet not 100% accurate. The reason is that we trained the model on the smaller dataset
and estimate the error on the validation dataset. This means, you get better idea
how the model is behaving on the unseen data.

Here we repeat the important information regarding your consent to the data collection. The following hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:

```{r, echo=FALSE}
textOutput("id_display2")
```
\
Please, keep it for your records in case you change your mind and want to delete your data.


