---
title: "Supervised methods - k-Nearest Neighbours"
output:
  html_document:
    df_print: paged
---

# Introduction
At the end of this lecture you will:

- know how to implement k-Nearest Neighbours algorithm (k-NN)
- be able to conduct ML experiment using `caret` package

> The lecture contains non-mandatory small task, which should help you to deepen 
your understanding. The are quoted as this text.

# Reproducibility
The k-NN is not influenced by the initialization problem like k-means algorithm, 
but when performing the ML experiment you will be splitting data into the folds.
Thus you will never get the same result. You need to "seed" properly random number
generator to obtain the stable and reproducible results. To do that you need to `set.seed` function
at the start of your code. This will guarantee that your results are stable 
between multiple runs. So, let us set the seed first:
```{r set-seed}
set.seed(2206)
```

# Required packages

We will be using `tidyverse` and `magrittr` libraries in this lecture.
In addition we will be using library `caret`. You will find out 
more about it later. Let us load them first:
```{r tidyverse}
library(tidyverse)
library(magrittr)
library(caret)
```

# Data
In this lecture we will be using data set called `iris`. You can read details of 
that dataset using `?iris`. It contains 150 samples of three different species
of iris and measurements of their sepal and petal leaf. We can explore the first 
few lines using `head`:
```{r head-iris}
head(iris)
```

As you can see we have 4 different features and one target variable, which represents
the iris species. At first we can explore, if there are missing variables in the 
data set:
```{r iris-missing-data}
iris %>% 
  summarise(across(everything(),~ sum(is.na(.))))
```
As the iris data set is the training data set you can see that there are no missing 
variables. But it is worth checking this at first to avoid future "problems". 
TO deal with missing values, you can explore imputation techniques (those are out
of the scope of this course, the most common one is replacing the missing values
with mean value of the column) We checked missing values and now we can observe 
the data set using the `ggplot2` package:
```{r iris-plot}
iris %>% 
  ggplot(aes(x = Sepal.Length, 
             y = Petal.Length,
             color = Species)) +
  geom_point()
```
You can observe that there is clear difference of the setosa class compared
to versicolor and virginica species. We can also observe that versicolor is somewhere
in the middle "bellow" the virginica, but the clear distinction is not possible.

> Try to play with the x- and y-axis features to see if the distribution of class
values is the same in all feature combinations.

# k-Nearest Neighbours

The k-Nearest Neighbors, which works with measuring the distance from new sample
to the samples in the training set. Since the algorithm did not involve any model
training we can proceed directly to the classification. In our case, we will create
an artificial scenario, on which we will develop the simple implementation of the
algorithm. The scenario involves removing one sample from the `iris` data set and
using it as new sample to classify. We will use the `dist` function to compute the
distance between the samples and then we will pick `k = 3` closest samples to the new
sample. The classification will be done by majority voting. 

At first, let us pick one random sample from `iris` dataset to play our new sample.
We will pick the samplenumber 104:
```{r knn-new-sample}
indx <- 104

new_sample <- 
  iris[indx,]
```

We also need to remove the sample from the training data set:
```{r knn-new-sample-remove}
training_data <- # remove sample from iris
  iris[-indx,]
```

Now, we have our new sample and we can compute distances between sample and all
other samples in `training_data` and pick `k = 3` closest samples. At first, we 
will append the sample to the training data set and drop the labels, which we do not 
want to use in the distance computation:
```{r knn-new-data}
new_data <- 
  new_sample %>% # add sample to the first position
  bind_rows(training_data) %>% 
  select(-Species) # drop labels
```

Next we will compute the distance between our new sample and all other samples in
the training data set. In addition, we will transform the distance matrix to the
standard matrix and extract the distances between our sample and the training data:
```{r knn-dist}
distances <-
  new_data %>%
  dist() %>% # count distance
  as.matrix() %>% 
  extract(1,) # extract distances 
```

Finally, we will pick the `k = 3` closest samples to our new sample:
```{r knn-pick-neighbours}
neighbours <-
  distances %>%
  order() %>% # return indexes of vector values sorted from min to max
  extract(2:4) # 3 elements (k=3) indx = 1 is our new sample
```

Now we know the neighbors of our sample, thus we can continue with the majority 
voting. We select the neighbors from the training data set and count their classes:
```{r knn-class}
classification <-
  training_data %>% 
  filter(row_number() %in% neighbours) %>% 
  count(Species) %>%
  arrange(desc(n)) %>% # sort by count, not necessary in this particular case
  extract(1,1) # extract the class label
```

The neighbours are all from the `virginica` class, thus our `new_sample` is 
classified as `virginica`. Since we know also the original label we can check,
if the algorithm have the classification correct:
```{r knn-check}
if(new_sample$Species == classification){
  print("Congratulations! You classified the sample correctly!")
} else {
  print("Incorrect!")
}
```
And it is correct! Congratulations, you implemented your own k-Nearest Neighbours
algorithm.

> Try to change the distance method for computation of distance in `dist` function. How the results changed?

> Try different number of `k`. What happens?

# Introduction to `caret` package and ML workflow

In previous parts you have learned how k-NN is working. But in reality you will 
be using the existing implementation of the algorithm. It is because you will never 
achieve the quality level of implementation compared to the widely used implementations. 
And of course, you will have no time to implement algorithms by yourself for every 
use case. Thus it is important to learn so-called Machine Learning workflow. In R we can 
make use of `caret` package, which provides the interface for many ML algorithm
implementations available in R. Plus it provides simplified way for conducting 
proper ML experiment. 

For the purpose of the demonstration we will be using the `iris` data set with just
two classes. To do that we will create new data set `iris_caret` containing only
classes `virginica` and `not virginica`:
```{r caret-iris}
iris_caret <- 
  iris %>% 
  mutate(Species = if_else(Species == "virginica", # is class virginica?
                           "virginica", # yes, keep it
                           "not virginica") # no, set the class to not virginica
         )
```

When you are performing the experiment with small data sets it is usually good 
approach to use cross-validation for estimating the method error. So for start 
we create 10 folds (by default), to which we will split the data:
```{r caret-folds}
# create folds for error estimation
folds <- createFolds(iris_caret$Species, list = FALSE)
```

The `folds` is the vector containing the number of corresponding fold to which 
the sample belongs. Next, we need to tell `caret` "engine" how it will be training
the algorithm. This is done by creating `trainControl` object containing our ML 
experiment configuration. In this case we will use 10-fold cross-validation:
```{r caret-traincontrol}
# caret control structure
fit_control <- trainControl(method = "cv", # train model using cross-validation
                            number = 10) # use 10 folds
```

For storing the results we will create empty `data.frame` containing the results:
```{r caret-results}
results <- data.frame(fold = numeric(0),
                      accuracy = numeric(0),
                      precision = numeric(0),
                      recall = numeric(0))
```

> Try to explore why using precision and recall is better than using only accuracy. 
Search the term "confusion matrix" and try to understand the concept. 

And now we can conduct our experiment:
```{r caret-ml-flow}
# for each fold
for(n in 1:max(folds)) {
  
  indx <- which(folds == n) 
  validation <- iris_caret[indx,] # use fold as a validation data set
  training <- iris_caret[-indx,] # use rest of data for training 
  
  # fit model
  fit <- train(Species ~ ., # predict Species using all (.) features
               method = "knn",
               data = training,
               trControl = fit_control)
  
  pred <- predict(fit, newdata = validation) # predict the class for validation dataset
  
  result <- confusionMatrix(data = pred, 
                            reference = as.factor(validation$Species), 
                            positive = "virginica") # obtain the measurements
  
  # store results
  results[n, ] <- c(n,
                    result$overall["Accuracy"],
                    result$byClass["Pos Pred Value"], # = Precision
                    result$byClass["Sensitivity"]) # = Recall
}
```

We first extract the training and testing data based on our initial call of `createFolds`.
Then we train the model using the `train` function. The `train` function is the main
function of the `caret` package, which is used to train the model. The function takes
the formula, method, data and `trainControl` object as the arguments. Formula specifies
the target variable and the features used for the prediction, method specifies 
the algorithm (in our case k-NN), data is the training data set and `trainControl`
is the object containing the configuration of the experiment. After the training
the model is used to predict the class of the validation data set. The predictions
are then used to compute the confusion matrix, which is used to compute the 
measures of performance. We set the positive class to `virginica`, that means that 
the `virginica` class is the class we are interested in and the precision and recall
values are computed for that class. The results are stored in the `results` data frame. 

You might notice that we are doing the cross-validation twice. In the outer loop
and withing the train function. This is because the `train` function is using the
cross-validation to estimate the error of the model and select the best model parameters
using the training data only. The outer loop is used to estimate the error of the model
on the validation data set.

> Try to explore the `train` function. What are the arguments of the function? 
What is the purpose of the `tuneGrid` argument?

Now we can analyse the results. We can compute the mean value of the measures,
which is the simplest approach to the analysis of the results. We can do that using:
```{r caret-final-results}
results %>% 
  summarise(across(2:4, mean))
```

We can observe that our k-Nearest Neighbours algorithm performed very well in classification 
of the `iris` data set. It achieved the accuracy of 0.97, precision of 0.95 and 
recall of 0.98. That means that 97% of all samples were classified correctly, 
95% of the samples classified as `virginica` were actually `virginica` and 98% 
of all `virginica` samples were classified correctly. In real world scenario, 
your work will end by providing the client with the final model trained on the 
whole data set and with  the detailed insights on how the model performs on the data.
This includes the measures of performance and how did you estimate them. 

> Try to experiment with caret results from one validation round? Can you find the 
final model produced for that round?

# Assigment

In today assignment we will be using [The HarvardX-MITx Person‐Course Dataset AY2013](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/26147). 
You can read more about this data set in the corresponding link. Your task is to 
predict if the student of course `HarvardX/CS50x/2012` will pass or fail the course
(`grade` is 0 for failed and 1 for passed) based on `ndays_act` and `nchapters` 
features using k-Nearest Neighbours (`knn`) classifier. To work on result prediction
you should be using `caret` package. 

## Instructions

- Create R Notebook named `09_submission.Rmd`, in which you will 
conduct the experiment.
- Download the data set from the provided [link](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/26147).
- Read data in R.
- Prepare the data set for the classification task.
  - Filter only students of the corresponding course `HarvardX/CS50x/2012`.
  - Filter students who do not have the grade value (`grade = NA`) and 
  focus on students who `explored` the course.
  - Select target variable and features.
  - Mutate target variable from 1/0 to pass/fail values.
  - Replace missing values of features (`NA`s) with 0.
- Load `caret`
- Fix the seed for reproducibility
- Split data set into 10-folds
- Define the `trainControl` object
  - Use 10-fold cross-validation for training.
- Fit model using training data and evaluate them on validation fold.
- Report the values of the Accuracy, Precision and Recall.

## Submission
Submit the task via Moodle in corresponding section. Your submission should contain
one `.Rmd` file named `09_submission.Rmd`.
