---
title: "Unsupervised Learning - assignment"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE, context="server"}
library(learnr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

library(readr)
library(tidyverse)
library(ggplot2)
library(caret)
library(patchwork)
```

## Assigment

In today assignment we will be using [The HarvardX-MITx Person‐Course Dataset AY2013](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/26147). 
You can read more about this data set in the corresponding link. Your task is to 
predict if the student of course `HarvardX/CS50x/2012` will pass or fail the course
(`grade` is 0 for failed and 1 for passed) based on `ndays_act` and `nchapters` 
features using k-Nearest Neighbours (`knn`) classifier. To work on result prediction
you should be using `caret` package. Data are already prepared for you in variable
`input_data`.

```{r exercise-setup, echo=FALSE}
input_data <- read_csv("data/assignment.csv")
```

> The work you are doing in chunks depends on previous one. Thus, if you get stuck
it might be a result of an error from previous steps.

## Check data   
Explore the `input_data` using `head`, `str` and `summary` functions.
```{r exercise-1, exercise=TRUE, exercise.setup = "exercise-setup"}
___(___)

___(___)

___(___)
```

## Filter course
At first, we need to `filter` students that are enrolled in the course `HarvardX/CS50x/2012`. 
Store the resulting data in variable `dataset`. Based on the `input_data` structure,
infere the column name that contains the course identifier.
```{r exercise-filter-course, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <-
  ___ |>
  ___(___ == "___")
```

## Filter students
Next, we will focus on students who have been graded and explored the course. That 
means we will `filter` out students who do not have a grade value (`grade = NA`) and
those who did not explore the course (`explored = 0`). Store the result again
in `dataset` variable and overwrite the previous value.
```{r exercise-filter-students, exercise=TRUE, exercise.setup = "exercise-filter-course"}
___ <-
  ___ |>
  ___(!___.___(___), 
      ___ == ___)
```

## Select target variable and features
Now, we have only samples (students) of our interest. Yet, we have lots of features
out of the scope of our task. We will `select` only the target variable `grade` and features
`ndays_act` and `nchapters`. Again, store the result in `dataset` variable and overwrite the previous value.
```{r exercise-select-target-features, exercise=TRUE, exercise.setup = "exercise-filter-students"}
___ <-
  ___ |>
  ___(___, ___, ___)
```

## Mutate target variable
The `grade` variable is currently represented as 0 (fail) and 1 (pass). We will `mutate` it to
pass/fail values. There is the function `if_else` that can be used to change the values of a variable based on a condition.
Store the result in `dataset` variable and overwrite the previous value.
```{r exercise-mutate-target, exercise=TRUE, exercise.setup = "exercise-select-target-features"}
___ <-
  ___ |>
  ___(___ = ___(___ == ___,
                "___",
                "___")
      )
```

## Replace missing values
Finally, we will replace missing values of features (`NA`s) with `0`. For that we will use `replace_na` function.
Store the result in `dataset` variable and overwrite the previous value.
```{r exercise-replace-missing, exercise=TRUE, exercise.setup = "exercise-mutate-target"}
___ <-
  ___ |>
  ___(list(___ = ___, 
           ___ = ___)
      )
```

## Setup classification task
Now we are ready to perform the classification task using k-Nearest Neighbours (`knn`) classifier.
At first, we need to fix seed (`1234`) and prepare the folds, training control object and empty results data frame (with 
`fold`, `accuracy`, `precision` and `recall` columns) like in our lecture. 
```{r exercise-setup-classification, exercise=TRUE, exercise.setup = "exercise-replace-missing"}
___(___)
# create (by default) 10 Folds based on grade variable of the dataset
folds <- ___(___$___, ___ = ___)

# caret control structure
fit_control <- ___(___ = "___", # train model using cross-validation
                   ___ = ___) # use 10 folds

# define empty results 
results <- ___(___ = ___(___),
               ___ = ___(___),
               ___ = ___(___),
               ___ = ___(___))
```

## Perform classification
Now we will perform the classification task using `knn` classifier using our 
well-known loop (like the one in the lecture).
```{r exercise-perform-classification, exercise=TRUE, exercise.setup = "exercise-setup-classification", exercise.lines = 30}
# for each fold
for(n in 1:max(___)) {
  
  indx <- ___(___ == ___) 
  validation <- ___[___,] # use fold as a validation data set
  training <- ___[-___,] # use rest of data for training 
  
  # fit model
  fit <- ___(___, # predict grade using all (.) features
             ___ = "___",
             ___ = ___,
             ___ = ___)
  
  predictions <- ___(___, ___ = ___) # predict the class for validation dataset
  
  cm <- ___(___ = ___, # predictions
            ___ = ___.___(___$___), # validation
            ___ = "___") # positive class
  
  # store results
  results[n, ] <- c(___,
                    ___$___["___"], # Accuracy
                    ___$___["___"], # = Precision
                    ___$___["___"]) # = Recall
}

results
```

## Show results
Finally, we will `summarise` the `results` and show the `mean` values of `accuracy`, `precision` and `recall`.
```{r exercise-show-results, exercise=TRUE, exercise.setup = "exercise-perform-classification"}
___ |>
  ___(___ = ___(___),
      ___ = ___(___),
      ___ = ___(___))
```

## Submission 
Now, you have the final result. Copy your code from the second to last step (*Perform classification*) and paste it 
to the Moodlelab in the [fourth assignment](https://moodlelab.hu-berlin.de/mod/assign/view.php?id=11434).
