---
title: "Introduction to R & RStudio - assignment"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---
```{r setup, include=FALSE, context="server"}
library(learnr)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)
```

## Assigment
In our first assignment we will be processing subset the ASSISTments dataset released
by Professor Neil Heffernan focusing on school year 2012/2013 (the dataset file is in
the zip). The ASSISTments is the online learning tool for deploying assignments with 
associated hints and resources online for their students. It is used by most of the 
Middle and High Schools in Worcester County, MA, USA.For us the data do not matter 
that much since we are going to create two functions: first will read the data from file, 
process them and outputs the results in other file; and the second will read the 
original data and produce histogram of selected column.

## I: read_transform_write

### Function 
Create function `read_transform_write`, which take one argument: `input`, which is a data.frame and returns it. Test the function call:
```{r exercise-setup, echo=FALSE}
input_data <- read.csv("data/assigment_data.csv")
```

```{r exercise-1, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___
}

read_transform_write(input_data)
```

### Select
Select the columns `user_id`,`assignment_id`,`start_time`,`end_time` and store them in new `data.frame` named `result`.
Return the `result`:
```{r exercise-2, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___ <- ___[, c(___)]
  ___
}

read_transform_write(input_data)
```
Hint: to delete column from `data.frame` you can assign `NULL` value to the corresponding column: `df$x <- NULL`

### New columns
Create new column `elapsed_time` in `result` containing time difference between `end_time` and `start_time`, round the difference to whole number; and column `message` in `result` containing string: `"The user <user_id> finished assignment <assignment_id> in <elapsed_time> seconds.`
```{r exercise-3, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___ <- ___[, c(___)]
  ___["___"] <- ___(___(___,___))
  ___$___ <- ___
  ___
}

read_transform_write(input_data)
```

Hints: 

  - it is possible to subtract two POSIXct values dt1-dt2 using `difftime` function, the result is in seconds
  - check function `round` 
  - for building the string you might search for the function `paste` or `paste0`
    
### Filter
The function should return only first 5 rows of the `result` data.frame. Adjust the code:
```{r exercise-4, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___){
  ___ <- ___[, c(___)]
  ___["___"] <- ___
  ___$___ <- ___
  ___(___)
}

read_transform_write(input_data)
```
Hint: there exists functions for examining structure of `data.frame` such as `str`, `head`, `tail`

## II: plot_histogram

### Function
Create function `plot_histogram`, which takes three arguments `df` data.frame, `col` string, `bins` number and returns `invisible(NULL)`:
```{r histogram-1, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___){
  ___
}

plot_histogram(input_data, "attempt_count", 3)
```

### Check column
Check that the `col` is a column in `df` and that it is numeric. If not, throw an error (you can test it with column names `attempt_count`, `type` and `non-existing-column`):
```{r histogram-2, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___) {
  if(!(___ %in% ___(___))){
    stop("Column does not exists")
  }
  if (!(___(___[[___]]))) {
    stop("Wrong column, values must be numeric")
  }
  ___
}

plot_histogram(input_data, "attempt_count", 3)
```
Hints: 
- you can throw error using `stop` function.
- to get names of the data.frame columns use `colnames` function

### Plot histogram
Now it is time to plot the histogram. Use `hist` function to plot the histogram of the column `col` with number of `bins`:
```{r histogram-3, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___) {
  if(!(___ %in% ___(___))){
    stop("Column does not exists")
  }
  if (!(___(___[[___]]))) {
    stop("Wrong column, values must be numeric")
  }
  ___(
    ___ = ___[[___]],
    ___ = ___
  ) 
}

plot_histogram(input_data, "attempt_count", 3)
```

### Pretty histogram
Finally, make the histogram pretty. Set the title of the histogram to `Histogram of column <col>`, set the color of the bars to `azure3` and set the x-axis label to the name of the column `col`:
```{r histogram-4, exercise=TRUE, exercise.setup = "exercise-setup"}
___ <- ___(___, ___, ___) {
  if(!(___ %in% ___(___))){
    stop("Column does not exists")
  }
  if (!(___(___[[___]]))) {
    stop("Wrong column, values must be numeric")
  }
  ___(
    ___ = ___[[___]],
    ___ = ___,
    ___ = ___,
    ___ = ___,
    ___ = ___
  ) 
}

plot_histogram(input_data, "attempt_count", 3)
```

## Submission 
**Now, you have the final result. Copy your code of `read_transform_write` and `plot_histogram` function and paste it 
to the Moodlelab in the [first assignment](https://moodlelab.hu-berlin.de/mod/assign/view.php?id=11431).**