---
title: "Introduction to R & RStudio"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---
```{r setup, include=FALSE, context="server"}
library(learnr)
library(glue)
library(dplyr)
library(curl)
library(jsonlite)
library(openssl)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

tutorial_options(exercise.eval = FALSE)

req <- curl_fetch_memory("https://api.ipify.org?format=json")

HASH <- as.character(md5(fromJSON(rawToChar(req$content))[[1]]))

event_recorder <- function(tutorial_id, 
                           tutorial_version,
                           user_id,
                           event, 
                           data) {
  REPO <- "/Users/jakubkuzilek/Work/learnr"
  
  if (file.exists(glue("{REPO}/data_{HASH}.rds"))) {
    t <- readRDS(glue("{REPO}/data_{HASH}.rds"))
  } else {
    t <- 
      tibble(
        time = .POSIXct(numeric(0)),
        tutorial_id = character(),
        tutorial_version = character(),
        user_id = character(),
        hash = character(),
        event = character(),
        data = list()
      )
  }
  
  t <- bind_rows(t, 
                 tibble(
                   time = Sys.time(),
                   tutorial_id = tutorial_id,
                   tutorial_version = tutorial_version,
                   user_id = user_id,
                   hash = HASH,
                   event = event,
                   data = list(data)
                 )
  )
  
  saveRDS(t, file = glue("{REPO}/data_{HASH}.rds"))
}

options(tutorial.event_recorder = event_recorder)
```

## Introduction

At the end of this tutorial you will:

- will know how to install R and RStudio
- know how to use RStudio IDE and RMarkdown notebooks
- understand R packages
- have insights into the base R including:
  - coding style
  - coding basics
  - data types
  - variables and functions
  - how to perform iteration using loops and apply function family
  - read and write contents of file
  - perform plotting using base R plotting functions
  
> For simplicity, we will not be using the R and RStudio installed on your computer. We will be
using interactive excercises created using [learnr](https://cran.r-project.org/web/packages/learnr/index.html) package.

> When you find `___` in the code chunk, it means that you need to fill in the missing part of the code.

## Informed Consent
This informed consent form is for a project investigating the interactions with the online
tutorial platform created using `learnr` package.

Name of Investigator: Dr. Jakub Kuzilek (jakub.kuzilek@hu-berlin.de)

Name of Supervisor: Prof. Dr. Niels Pinkwart (pinkwart@hu-berlin.de)

Name of Institution: Humboldt-Universität zu Berlin

**You can download this informed consent in the Moodle course.**

**The data from multiple tutorials will be collected and analysed.The informed consent is the same. For each session, you will get hash, which is the only unique identifier of your interactions with the material.** 

### Information sheet

#### Introduction of investigators
Jakub Kuzilek is a researcher in fields of educational data mining and learning analytics since completion of his PhD in Artificial Intelligence and Biocybernetics at the Czech Technical University. In 2020 he joined Prof. Dr. Pinkwarts's research group and currently working on the project IMPACT.

#### Purpose of the study
The purpose of this study is to investigate the interactions with the online tutorial platform created using the `learnr` package. The data will be used to improve the design of the tutorial and to understand how users interact with the platform.

#### Type of Research Intervention
To support and investigate these learning and interaction processes interaction data will be used. In addition performance data of the students will be collected.

#### Research Data Management
The collected data will be anonymized and merged. The collected data are stored on the servers of the Computer and Media Service (CMS) of the Humboldt-Universität zu Berlin in Germany and the computers of the researchers. The data access is protected and regulated via a login and the access is restricted to the members of the project team. The personal data will be stored on secured location accessible via authenticated encrypted connection (VPN and SSH). The data are archived for another 10 years after the end of the project according to good scientific practice. The anonymized data may be published to enable Open Access for the general public. The participant may gain access to his/her recorded data, request a correction to his/her personal data record, or withdraw from the research study by contacting the Investigator:
Jakub Kuzilek at jakub.kuzilek@hu-berlin.de

In case of doubt as to the lawfulness of the processing of personal data, data subjects have the right to lodge a complaint with a Member State supervisory authority. The competent supervisory authority for Humboldt-Universität zu Berlin is Die Berliner Beauftragte für Datenschutz und Informationsfreiheit, Friedrichstraße 219, 10969 Berlin. The official data protection officer for Humboldt-Universität zu Berlin is: Gesine Hoffmann-Holland, Tel: +49 (30) 2093-20022, datenschutz@uv.hu-berlin.de.

#### Voluntary participation
Participation in this research project is entirely voluntary. It is up to the participant to decide whether or not to participate and there is no explicit or implicit obligation to participate. The participant has the possibility to cancel the participation at any time - even after the signing of the Informed Consent - without any disadvantages for the participant. 

#### Documentation of the consent
One copy of this document will be kept separately from the research records of this study using a pseudonym id in order to remove data after the request from the participant. The contact details of the participants will be deleted after the project is finished. To assure data deletion of the requesting subject a unique identifier in form of the email address and is required.

### Certificate of consent
- I agree to the collection of my pseudonymized interaction data in the course "Machine Learning for Education".
- I agree to the collection of my pseudonymized perfomance data in the course "Machine Learning for Education".
- I agree to have the anonymized interaction data and the anonymized performance data published under the Creative Commons Attribution 4.0 International (CC-BY) Licence (https://creativecommons.org/licenses/ by/4.0/) and so I allow it to be used by other people for other projects. 
- I agree to have the interaction data, and the performance data archived for 10 years after finishing of the project.

This hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:
**`r HASH`**

Please, keep it for your records in case you change your mind and want to delete your data.

```{r informed-consent, echo=FALSE}
question_radio("I agree to the collection of my pseudonymized interaction data in the course 'Machine Learning for Education'.", 
               correct = NULL, 
               answer("yes", correct = TRUE),
               answer("no", correct = TRUE)
)
```
## Installing the R and RStudio

R is both programming language and interactive environment for statistical computing. 
That means after installation of the R distribution you can start coding/doing data
analysis without any other software installed. It is open-source implementation of
language S developed in Bell Lab by John Chambers in 1975. The R has been released 
around 1993 by Robert Gentleman. RStudio IDE project started around 2010 and it is 
developed since then. Posit (RStudio) Co. takes care about the development and also
it is one of the contributor to the R community.

### Installing R

To install R you need to visit [CRAN (Comprehensive R Archive Network)](https://cran.r-project.org/), 
which makes available all versions of R and its packages. There exists distributions 
for [Linux](https://cran.r-project.org/bin/linux/), [macOs](https://cran.r-project.org/bin/macosx/) 
and [Windows](https://cran.r-project.org/bin/windows/).

### Installing RStudio

After you install the R you can install the [RStudio Desktop](https://posit.co/download/rstudio-desktop/),
which provides the nice IDE for R. 

### Using R, RStudio & R notebooks

When you start the RStudio, you will enter the main screen of the IDE. The bottom left 
tab contains started instance of the R interactive console. On the right side there 
are on top several tabs with the information about the Environment variables,
Command History and other tabs. At the bottom there is file browser tab, which will be 
the most important now.

### Using R interactive environment

As it has been said R is also interactive programming environment. That means you
can start typing the commands in the console at the bottom left tab. We will speak
about the commands and use of R in next part. For now, two not so obvious commands:
`?` and `??`, which in combination with name of the function will provide you with the
manual for that function. Try it now (we will also test the learnr interactive environment):

```{r help-exercise, exercise=TRUE}
___length
```

### Using R scripts

R does a lot more than just interactive programming. It is also a full-fledged programming
language. You can write scripts and functions. In our case R Script and R Notebook file typesare of most
importance. R Script is textual file ending with `.R` containing executable commands and/or
function definitions. R Notebook provides the "Jupyter" like functionality for R.
The `.R` files needs to be "sourced" into the R via `source` function to be able to 
execute code in them. 

### R notebooks

As mentioned before R Notebook provides "Jupyter" like functionality to RStudio IDE. This
is very handy especially in case when one performs data analysis, which takes lot of time.
The R Notebooks can be structured as a document using RMarkdown mutation of Markdown language. 
In addition to the text, you can insert code chunks containing parts of your analysis.
Data is share between chunks of the document so you can reuse variables you computed earlier in 
your document. When you first try to start R notebook on fresh installed RStudio
the message asking you about installing additional packages pop-up.

### Recommended coding style

The recommended coding style for our lectures is the one developed by Hadley Wickham and
it is called [tidyverse style guide](https://style.tidyverse.org/index.html). In 
principle try to use lower_snake_case for your variable naming, `"` for string quoting,
and `<-` for variable assignment instead of `=`. The arrow is "special" for R, 
but is established well in the language.

## R packages \& CRAN

R itself is very powerful for data analysis and statistical testing, you would be 
surprised how much you can do without any extension. However, over the years the 
mature and well formalized system of packages evolved to support the users with 
additional functions. The R use package system, which is maintained by group of developers.
It guarantees the users certain level of quality through the package review system
on package submission and the system takes care about the dependencies between packages.
All R packages are available on CRAN. In addition to CRAN there exists other repositories 
such as Bioconductor, which are "clones" of CRAN and provides additional packages
for certain research fields. You can also use packages available for example on GitHub, but 
you need to be aware that those does not go through any review process and might 
not have clear author/maintainer.

In R you can install any package using:

```{r install-packages, eval=FALSE}
install.packages("<name-of-package>")
```

If you want to install more packages at once:

```{r install-many-packages, eval=FALSE}
install.packages(c("<name-of-package-1>", "<name-of-package-2>", "<name-of-package-3>", "<name-of-package-4>"))
```

```{r dplyr-setup, echo=FALSE}
chooseCRANmirror(ind = 38)
```

Try to install the package `dplyr` (when you succeed you will see the information about saving the data):
```{r dplyr-exercise, exercise=TRUE, exercise.setup = "dplyr-setup"}
# Install package dplyr
install.packages("___")
```

To list all packages installed in your computer just run following command:
```{r list-packages-exercise, exercise=TRUE}
# Try to list all installed packages within your distribution of R
installed.packages()
```

To use installed package in the R environment you need to call `library` function to 
load the package.

```{r, eval=FALSE}
library(<name-of-package>)
```

## Basics of R programming

You can use R as any other programming language for performing basic math operations. 
Try to add two numbers:

```{r addition, exercise=TRUE}
___ + ___
```

You might notice that the result is produced on line starting with `[1]`. This
is the way how the R is telling you the index of the first item on line. In such a 
way the analysis of vectorized results is much simpler. Check this behaviour in the 
following example:

```{r vector, exercise=TRUE}
1 : 100
```

For more details about the basic operations in R you can check the for example [A Short Introduction to R Programming](https://bookdown.org/manishpatwal/bookdown-demo/basic-operations-in-r.html).

## Data types

In R there are several data types, which are completely natural to any other programming
language. To explore what type is the value or variable you can use `class()` function.
The numeric values can be stored as:

```{r numerics, eval = FALSE}
# integer
1L

# double ~ numeric
1

1.0

# complex
2+3i
```

The numeric type is default for the numerical values. Characters and strings are represented by data type
`character` and you can use `'` or `"` quotes for defining them:

```{r strings, eval = FALSE}
"this_is_string" # recommended
'this_is_other_string'
```

Try to create a string containing `"` inside it. There are two possible ways (one involving `\` and the second involving `'`):
```{r string-quotes, exercise=TRUE}
___A string with ___" inside it___
```

Logical values are then expressed using:

```{r logical, eval = FALSE}
TRUE # recommended
FALSE # recommended
# or
T
F
```

The final data types we will be speaking about are dates and times. Dates are represented
as number of days between the date and 1970-01-01. To create date from the string
you can use function `as.Date()`. When one needs to handle time the most appropriate 
way is to use POSIX classes, which stores the datetime as list of its components
- hour, min, sec, mon,etc. Let us create some dates and datetimes:

```{r dates, exercise=TRUE}
as.Date("___-___-___")

as.POSIXct("___-___-___ ___:___:___")
```

## Data structures

### Vector
Since R is oriented on data analysis and manipulation it comes naturally with several
data structures. The simplest type is a vector containing values of the same type. 

To create an empty vector you can use function `vector`. Try it now: 
```{r vectors, exercise=TRUE}
# create empty vector of numbers
___("numeric", 4)
```

For creating vector of values you can use several different ways:
1. `seq()` function
2. `c()` function
3. `rep()` function
4. `:` operator

Try  to create vector of values from 1 to 10 using `seq()` function:
```{r seq-function, exercise=TRUE}
seq(___, ___, ___) # from, to, by
```

The same vector can be created using the `:` operator. Try it now:
```{r colon-function, exercise=TRUE}
___ : ___ : ___
```

The `rep()` function is used for repeating the values in the vector:
```{r rep-function, eval=FALSE}
rep(1:3, times = 3) # repeat the whole vector 3 times
```

And finally the `c()` function is used for creating vector of specific values:
```{r c-function, eval=FALSE}
c(1, 4, 5, 7)
```

### Matrix and array

The "extensions" of vector to more dimensions are matrix (2D):

```{r matrix}
matrix(1 : 10, nrow = 2, ncol = 5)
```

And array (>2D):
```{r array}
array(1 : 12, dim = c(2, 2, 3))
```

Create the matrix of 4 rows and 5 columns with values from 1 to 20:
```{r matrix-exercise, exercise=TRUE}
matrix(___ : ___, ___ = 2)
```

### List

In addition to listed previously the `list` data structure exist. It can store values
of different data types:

```{r list}
list(1, 2, "Hello", c(1, 2))
```

The list print out is specific at first list index is printed with `[[1]]` and then 
the corresponding item from the list is printed out using standardized print output. 

### Data frame
Finally, to work with tabular data the `data.frame` data structure exists. On background
it works as a list of lists with specific requirements on same number of items in
each sub-list:

```{r dataframe}
data.frame(first_col = c(1, 2, 3), second_col = c("a", "b", "c"))
```

The `data.frame` requires you to name each column of the data by name so it can be 
distinguished from other columns.

### Factors 

Last but not least is special data structure for categorical variables called `factor`.
It is special kind of vector, which works with the "labelled" character values. It 
provides faster processing of the categorical data and it can also check the ordering for
ordinal categorical variables. Let us see examples:

```{r factors, eval=FALSE}
factor(c("dog", "cat", "dog", "dog", "cat"))

factor(c("A", "B", "D"), levels=c("A", "B", "C", "D"), ordered = TRUE)
```

Create `factor` with values `"A", "B", "E"` but levels `"A", "B", "C", "D"` and 
observe the result:
```{r factor-exercise, exercise=TRUE}
factor(c(___), levels=c(___), ordered = TRUE)
```

## Type/structure conversion

For converting types or structures to another type you can use commands starting with `as.`:

```{r conversion, eval=FALSE}
as.character(1)

as.numeric("1")

as.list(seq(1, 10, 2))

as.vector(factor(c("A", "B", "D")))

as.numeric(factor(c("A", "B", "D")))
```

You can notice that factor can be converted in two ways - as a character/string vector
or numeric vector. The second one contains the underlying level numeric values.

Try to convert the factor `"A", "B", "D"` to character and numeric vector:
```{r conversion-exercise, exercise=TRUE}
as.___(factor(c(___)))
```

## Objects

### Variables
Now that we know the basic of values and data structures we can make use of objects, 
which is basically name upon which call the stored data are accessible. To assign data 
to object use `<-` or `=`: 

```{r objects}
x <- 1 # assign numeric value to object x

y = 2 # assign numeric value to object y

die <- 1 : 6 # assign vector to object die
```

Please, note that the `=` sign for value assignment is not recommended way. The reason
is simple it has not been in the original specification and it has been added later thus
standard syntax is biased towards preferring the arrow.

You can work with the objects like with the values. Try to divide the variable `die` by `2`:
```{r object-operations-setup}
die <- 1 : 6 # assign vector to object die
```

```{r object-operations, exercise=TRUE, exercise.setup = "object-operations-setup"}
die ___ ___
```

In addition to the basic operations you can also use the `%*%` and `%o%` operators for inner and outer multiplication
of two vectors. Try to use them on the `die` object:
```{r object-operations-2, exercise=TRUE, exercise.setup = "object-operations-setup"}
die ___ die 
```

### Named vectors, list and data.frames
Vector and list items can be named. This can be done via `names` function. Name the 
elements of the vector `die`. The names should be `"one", "two", "three", "four", "five", "six"`:

```{r named-vectors, exercise=TRUE, exercise.setup = "object-operations-setup"}
names(die) <- c(___)
```

## Accessing items in structures

### Numerical indexing 

When dealing with vectors, matrices and others structures it is crucial to understand how to access the
individual items of corresponding structure. R provides multiple options to access the items.
At first, you can use a numeric index. To extract one single item, use `[]` brackets with the index of the item you want to extract:
```{r numeric-indexing-single}
die[1]
```
To extract multiple items, you can use a vector of indices:
```{r numeric-indexing-multiple}
die[1 : 3]
```
Or you can use a vector of indices to extract non-consecutive items:
```{r numeric-indexing-multiple-2}
die[c(1, 3, 6)]
```

When dealing with matrices you can use the same indexing, but you need to specify the row and column index.
```{r matrix-indexing}
(example_matrix <- matrix(1 : 6, nrow = 2, ncol = 3))
```
The outer brackets `()` are used for forcing R to print expression result in the console.
Now you can access the items in the matrix using the row and column index:
```{r matrix-indexing-2}
example_matrix[1, 2]
```
You can also extract the whole row or column:
```{r matrix-indexing-3}
example_matrix[1, ]
```

When dealing with lists and data.frames the indexing is "little bit" more tricky. 
We will first define the list: 
```{r list-indexing}
(example_list <- list(1, 2, "Hello", c(1, 2)))
```
And now we will extract the items from the list. The first way is to use `[]` brackets,
which extracts the "sub-list" from the list:
```{r list-indexing-2}
example_list[4]
```
To obtain the item(s) in the "sub-list" you need to use `[[ ]]` brackets:
```{r list-indexing-3}
example_list[[4]] 
```

And of course you can chain the indexing. You can extract 2nd item from the 4th "sub-list":
```{r chain-indexing-setup, echo=FALSE}
example_list <- list(1, 2, "Hello", c(1, 2))
example_matrix <- matrix(1 : 6, nrow = 2, ncol = 3)
```

```{r chain-indexing}
example_list[[4]][2] # returns 2nd item of the item in the 4th "sub-list"
```

Let us have some practice. We will be working with our `die` vector and `example_list`. 
Try to first extract every second item from the `die` vector:
```{r die-indexing-setup, echo=FALSE}
die <- 1 : 6
example_list <- list(1, 2, "Hello", c(1, 2))
```

```{r die-indexing, exercise=TRUE, exercise.setup = "chain-indexing-setup"}
die[___]
```

Now try to extract the first and last item values from the `example_list`:
```{r example-list-indexing, exercise=TRUE, exercise.setup = "chain-indexing-setup"}
example_list[___]
```

### Logical indexing 
Other option is to use logical indexing where you define logical "vector" of the 
same dimension as the structure you would like to use for item extraction:

```{r logical-indexing}
logical_indx <- c(TRUE, FALSE, FALSE, TRUE, FALSE, FALSE)

die[logical_indx]
```

Try to change logical_index to extract only the even numbers from the `die` vector:
```{r logical-indexing-exercise, exercise=TRUE, exercise.setup = "chain-indexing-setup"}
logical_indx <- c(___, ___, ___, ___, ___, ___)

___[___]
```

### Named indexing
If the structures you are using have named items then you can access items by their name. 
For example, for a vector `die` with names `"one", "two", "three", "four", "five", "six"` you can access the items using the name.
```{r name-indexing-setup, echo=FALSE}
die <- 1:6
names(die) <- c("one", "two", "three", "four", "five", "six")
```

There is difference between using `[]` and `[[]]` brackets. The first one returns the item and its name: 
```{r name-indexing}
die["one"]
```

And the second one returns the item itself:
```{r name-indexing-1}
die[["one"]]
```

We can use the column names for accessing values of `data.frame`. For example, if we have a data.frame `example_data_frame` with columns `first_col`, `second_col`, and `third_col`:
```{r name-indexing-2}
(example_data_frame <- data.frame(first_col = c(1, 2, 3), 
                                 second_col = c("a", "b", "c"),
                                 third_col = seq(5,15,5)
                                 ))
```

We can access the columns using the column names:
```{r name-indexing-3}
example_data_frame[["first_col"]]
```

Try to experiment with single and double brackets:
```{r name-indexing-4, exercise=TRUE}
example_data_frame___"first_col"___ 
```
When we select multiple data.frame columns, we create a new data.frame:
```{r name-indexing-5}
example_data_frame[c("first_col","second_col")] 
```

### `$` "indexing"
Finally, the `list` and `data.frame` items can be accessed via special `$` notation:
```{r dollar-indexing-setup, echo=FALSE}
example_data_frame <- data.frame(first_col = c(1, 2, 3), 
                                 second_col = c("a", "b", "c"),
                                 third_col = seq(5,15,5)
                                 )
```

```{r dollar-indexing}
example_data_frame$second_col
```

You can also use the `$` notation for adding the columns to the `data.frame`. Try it now:
```{r add-column, exercise=TRUE, exercise.setup = "dollar-indexing-setup"}
example_data_frame$___ <- 1

# print out the result
example_data_frame
```

## Conditional statements 

For conditional execution of commands you can use `if` statements:
```{r if}
day <- TRUE
if(day) {
  print("It is day!")
}
```

You can, of course, extend the statement with else part:
```{r if-else}
day <- FALSE
if(day) {
  print("It is day!")
} else {
  print("It is night!")
}
```

And of course you can chain the if-else with else if blocks:

```{r if-else-if}
height <- 160

if(height < 130) {
  print("You are small!")
} else if(height < 170) {
  print("You are medium!")
} else if(height < 200) {
  print("You are large!")
} else {
  print("You are extra large!")
}
```

## Loops

To perform repetitive task one can employ loops. In R the loops are not efficiently
implemented and they should be used in experimental setting, when the complexity 
of the problem is too large or you need something to run indefinetely (while loop).

### for loop

The for loop iterate over the sequence of numbers defined in the `()`:
```{r for-loop}
for(m in 1:5) {
  print(m)
}
```
Try to adjust the for-loop to print out only numbers "1", "3", "8" and "15":
```{r for-loop-exercise, exercise=TRUE}
# with numeric vector
for(m in ___) {
  print(m)
}
```

### while loop

The while executes the code within its body until the condition in `()` is satisfied:

```{r while-loop}
counter <- 0

while(counter <= 5) {
  counter <- counter + 1
  print(counter)
}
```

## Functions

To efficiently use and re-use your code you will be using functions. You already
met several functions provided by base R such as `length`, `names` or `install.packages`. 
But the real fun starts with own function. To create a function one needs to 
specify its arguments in "header" and what should happen with them in function "body". 
```{r function, eval=FALSE}
<function-name> <- function(<arguments>) {
  <function-body>
}
```
Function automatically returns the result of the last expression. Let us show example.
We first define the function `roll_dice`, which simulates rolling a die with `number_of_sides` sides `how_many_times` times:
```{r simple-function}
roll_dice <- function(number_of_sides, how_many_times){
  sample(1:number_of_sides, how_many_times, replace = TRUE)
}
```

Next we need to call the function with the arguments:
```{r simple-function-execute}
roll_dice(6, 20)
```

Sometimes it is handy to define default arguments values for the function call. To 
do that you need to specify them in the function "header". Try specify the arguments 
default values for the function `roll_dice_extended` and then call it without 
specifying the arguments:
```{r simple-function-default, exercise=TRUE}
roll_dice_extended <- function(number_of_sides = ___,
                               how_many_times = ___){
  sample(1:number_of_sides, how_many_times, replace = TRUE)
}

___()
```

The R is checking the arguments of the function and assigns them according to their
order in function call. However, you can specify the argument by using its name without
necessity to specify all other arguments. So if our default values from previous 
example are set to `number_of_sides = 6` and `how_many_times = 10`, you can call the function
with only the second argument:
```{r simple-function-default-execute-second-arg-setup, echo=FALSE}
roll_dice_extended <- function(number_of_sides = 6,
                               how_many_times = 10){
  sample(1:number_of_sides, how_many_times, replace = TRUE)
}
```

```{r simple-function-default-execute-second-arg}
roll_dice_extended(how_many_times = 20)
```

## Apply family of functions
As we said before the loops in R are inefficient. So how one can perform the efficient
iteration operations over the vectors and lists. For that purpose the R have the
system of "apply" functions family, which applies the selected function to the 
defined object. In addition to better performance apply-functions also lead less 
errors in the code. There exists multiple functions but for us the most important
functions would be `lapply` (aka list-apply) and `sapply` (simplify-apply). The
`lapply` function takes the list or vector of values and applies the selected function
on each of them producing the list of results:
```{r lapply}
square <- function(number) {
  number^2
}

lapply(seq(1,4,1), square)
```

The `lapply` is useful when the produced result is complex and the structure needs 
to be stored in list. However, in the most cases one needs to perform simple
operation and get simplified result in form of vector. And that is the point where 
`sapply` comes in play:
```{r sapply}
sapply(seq(1,4,1), square)
```
There exist other functions from the family: `apply`, `tapply`, `vapply` and `mapply`, 
but those are out of scope of this introduction.

Try to use the function `sapply` to calculate the print the result squaring of the numeric vector. Try on vector 1 to 10:
```{r sapply-exercise, exercise=TRUE}
___ <- function(___) {
  result <- ___^2
  ___(result)
}

sapply(___, ___)
```

Notice that `sapply` returns the result anyway. That is the nature of the function, which
automatically assigns the individual results to items of vector or list.

## Reading and writing (csv) files
R provides set of functions for writing and reading data into the environment. We will
be using functions for `csv` files manipulation. To store data in file you can use
function `write.csv`:
```{r write.csv, eval=FALSE}
write.csv(<data.frame>, "<file-name>")
```

It takes multiple arguments, but most of the time you need to specify the first (data) and
the second one (path). And we can recover this stored dataset using `read.csv` function:
```{r read.csv, eval=FALSE}
<variable-name> <- read.csv("<file-name>")
```

## Base plotting
The base R gives you the opportunity to visualise data you are working with. In our
introduction we will be working with two base plots: scatter plot (`plot` function)
and histogram (`hist` function).

### scatter plot
Scatter plot enables you to explore the dependencies between two features in your 
dataset. Each sample of your data is represented as a point in two dimensional 
plane with x and y values based on the selected features. We will use famouse `iris` dataset
in our examples. We can plot relationship between Petal Length and Width features:
```{r scatter-plot}
plot(iris$Petal.Length, iris$Petal.Width)
```

The function takes as the first two arguments the vectors of values for x and 
y axis and the plots them. Our `iris` dataset is stored as `data.frame` thus we 
can access the values of Petal Length and Width using `$` notation, which extracts
the vectors out of the `data.frame` structure. The plot function comes with additional
arguments, which give you opportunity to style the plot. Let us extend our example by adding
Some features such as labels and title. Try it now:
```{r extended-scatter-plot, exercise=TRUE}
plot(iris$Petal.Length, iris$Petal.Width, 
     xlab = "___",
     ylab = "___",
     main = "___",
     pch = 20)
```

The arguments changes the textutal information in the labels and adds title of the plot. With 
the last argument we changed points from circle to dots (`pch` gives you the possibility
to specify the plotting symbol).

### Histogram
Often it is necessary to explore the distribution (or its approximation) of one feature in 
dataset. To do that histogram function comes in play:
```{r histogram}
hist(iris$Petal.Length)
```

The `hist` function produce the bar chart or histogram of values in the provided
vector with x axis depicting the possible values and y axis showing their frequency
in dataset. Again, we can adjust the produced figure by adding some labels and title:
```{r extended-histogram, excercise=TRUE}
hist(iris$Petal.Length,
  main="___",
  xlab="___",
  col="darkblue",
  freq=FALSE
)
```

The produced histogram contains our specified label and title. We also changed
the color of bars and instead of the frequency we decided to show relative frequency
(aka density) of values (by setting `freq` argument to `false`).

## Conclusion
The base plotting concludes our introduction to R. Next comes the first assignment,
which will available after of our seminar.

Here we repeat the important information regarding your consent to the data collection. The following hash will be used to identify you in the data from this lecture and to delete your data after the project is finished:
**`r HASH`**

Please, keep it for your records in case you change your mind and want to delete your data.


